\documentclass{article}
\usepackage{hyperref}
\usepackage{graphicx} % Required for inserting images
\usepackage{tikz}
\usepackage{soul}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{float}
\usepackage{parskip}
\usepackage{pgfplots}
\usepackage{xcolor}
\usepackage{mdframed}
\usepackage[margin=1.0in]{geometry}
\usepackage[shortlabels]{enumitem}

\newcommand*{\Perm}[2]{{}^{#1}\!P_{#2}}%
\newcommand*{\Comb}[2]{{}^{#1}C_{#2}}%
\newcommand*{\qed}{\hfill$\square$}%
\newcommand*{\txt}[1]{\text{ #1 }}%
\newcommand*{\iprod}[1]{\langle #1 \rangle}
\newcommand*{\fora}{\txt{}\forall}%
\newcommand*{\rr}{\mathbb{R}}%
\newcommand*{\zz}{\mathbb{Z}}%
\newcommand*{\bigo}{\mathcal{O}}%


\newcommand*{\simu}{\sim_{\mathcal{U}}}
\newcommand*{\dotslist}[1]{#1_{1},\dots, #1_n}

\newlength\tindent
\setlength{\tindent}{\parindent}
\setlength{\parindent}{0pt}
\renewcommand{\indent}{\hspace*{\tindent}}

\newmdenv[
    topline=false,
    bottomline=false,
    rightline=false,
    backgroundcolor=gray!5,
    linecolor=blue!100,
    skipabove=\topsep,
    skipbelow=\topsep
]{siderules}

\newenvironment{aside}{\begin{siderules}\setlength{\leftskip}{0.3cm}}{\end{siderules}\setlength{\leftskip}{0pt}}

\title{MATH406 Notes}
\begin{document}
\author{Nyx}
\maketitle

\parskip = 1em
\pgfplotsset{compat=1.17}
\setcounter{section}{0}

\section{Intro}

Number theory is mostly about the properties of integers 

We will preemptively assume the entirety of set theory and the basics of lin alg. 

Textbook: Divisibility

\section{Lecture 1}

\subsection{2.1 - Divisibility}

We can add/subtract/multiply integers to get integers (eg prop of field, closed).

\textbf{Definition:} Let $a$, $b$ be integers, $b \neq 0$. We say that $b$ divides $a$ (denoted $b|a$) if there exists an integer $k$ such that $a = bk$.

Additionally, 
\begin{itemize}
    \item If $b$ does not divide $a$, we write $b \nmid a$.
    \item When $d|a$, we say that $a$ is a multiple of $d$ and that $d$ is a divisor (or factor) of $a$.
    \item For any nonzero integer $d$ we have $d|0$ since $0 = d \cdot 0$.
    \item We never consider division by zero, so $0|a$ is not defined.
\end{itemize}

\underline{Convention:} When we talk about the set of divisors of a positive integer, we mean the set of positive divisors. For instance, the divisors of 6 are 1,2,3,6 (not -1,-2,-3,-6).

\textbf{Proposition:} Let $a,b,c$ integers. Then $a|b$ and $b|c$ implies $a|c$.

\textbf{Proof:} Since $a|b$, there exists an integer $k$ such that $b = ak$. Since $b|c$, there exists an integer $l$ such that $c = bl$. Thus, \begin{align*}
    c &= bl \\
      &= (ak)l \\
      &= a(kl)
\end{align*}
Since $kl$ is an integer under field closure, we have $a|c$. \qed

\textbf{Definition:} A linear combination of integers is an expression of the form $ax + by$ where $x,y$ are integers.

"A divisor of both $a$ and $b$ is a divisor of any linear combination of $a$ and $b$."

\textbf{Proposition:} Bilinearity of division; that is, $d|ax + by$ if $d|a$ and $d|b$.

\textbf{Proof:} Since $d|a$, there exists an integer $k$ such that $a = dk$. Since $d|b$, there exists an integer $l$ such that $b = dl$. Thus, \begin{align*}
    ax + by &= (dk)x + (dl)y \\
            &= d(kx) + d(ly) \\
            &= d(kx + ly)
\end{align*}
Since $kx + ly$ is an integer under field closure, we have $d|ax + by$. \qed

\textbf{Corollary:} Suppose $a,b,d$ are integers. Then, if $d|a$ and $d|b$ then $d|(a-b)$ and $d|(a+b)$.

\textbf{Proof:} Let $x = 1$ and $y = -1$ in the previous proposition to get $d|(a-b)$. Let $x = 1$ and $y = 1$ to get $d|(a+b)$. \qed

\subsection{2.2 - Euclid's Theorem}

\textbf{Definition:} A \underline{prime number} is an integer $p\geq 2$ whose only positive divisors are 1 and $p$ itself. An integer $n \geq 2$ that is not prime is called \underline{composite}. 

\textbf{Theorem (Euclid's Theorem):} There are infinitely many prime numbers.

\section{Lecture 2}

\textbf{Proof of Euclid's Theorem:} Suppose there are finitely many primes $p_1, p_2, \ldots, p_n$. Let \begin{align*}
    Q &= p_1 p_2 p_3 \cdots p_n + 1
\end{align*}

Then $Q$ is either prime or composite. If $Q$ is prime, then $Q$ is a prime not in our list. If $Q$ is composite, then $Q$ has a prime divisor $p$. Since $p$ divides the product $p_1 p_2 \cdots p_n$, it follows that $p$ does not divide $Q - p_1 p_2 \cdots p_n = 1$. Thus, $p$ is not in our list of all primes. In either case, we have a contradiction. Therefore, there are infinitely many primes. \qed

\subsection{2.4 - The Sieve of Eratosthenes}

\textbf{Theorem (Sieve of Eratosthenes):} To find all primes less than or equal to a given integer $n \geq 2$, we can use the following algorithm:
\begin{enumerate}
    \item Write down all integers from 2 to $n$.
    \item Start with the first number $p$ in the list (which is 2). Circle $p$ and cross out all multiples of $p$ greater than $p$.
    \item Find the next number in the list that is not crossed out. If there is no such number, stop. Otherwise, let $p$ be this new number and repeat step 2.
    \item When the algorithm stops, the circled numbers are all the primes less than or equal to $n$.
\end{enumerate}

\textbf{Proof:} Let $p$ be any prime less than or equal to $n$. When we reach step 2 with this value of $p$, we circle it since it has not been crossed out (as it is prime). Thus, all primes less than or equal to $n$ are circled when the algorithm stops. \qed

\section{Lecture 3}

\subsubsection{2.5 - The Division Algorithm}

Consider the grade school long division problem. We want to do, for instance, $95$ divided by $7$. We can write \begin{align*}
    95 &= 7 \cdot 13 + 4
\end{align*}

\textbf{Theorem (The Division Algorithm):} Let $a,b\in \mathbb{Z}$ and $b>0$. Then $\exists! q,r\in \mathbb{Z}$ st \begin{align*}
    a &= bq + r \txt{ with } 0 \leq r < b
\end{align*}

Note that if $b>a$ then $r=a$ and $q=0$. 

\textbf{Proof:} (Existence) Let $q$ be the largest integer such \begin{align*}
    q\leq \frac{a}{b}<q+1
\end{align*}

Then $bq\leq a<bq+b$. and so $0\leq a-bq<b$. Let $r=a-bq$. Then $a=bq+r$ with $0\leq r < b$. 

(Uniqueness) Suppose there exist $q',r'$ such that \begin{align*}
    a &= bq' + r' \txt{ with } 0 \leq r' < b
\end{align*}
Then \begin{align*}
    bq + r &= bq' + r' \\
    b(q-q') &= r' - r
\end{align*}
If $q \neq q'$, then $|b(q-q')| \geq b$ since $|q-q'| \geq 1$. But $|r'-r| < b$ since both $r$ and $r'$ are between $0$ and $b$. This is a contradiction. Thus, $q=q'$, which implies $r=r'$. \qed

\textbf{Example:} \begin{enumerate}
    \item For $a=100, b=9$, we have $100=9\cdot 11 + 1$.
    \item For $a=-100, b=9$, we have $-100=9\cdot (-12) + 8$.
\end{enumerate}

\textbf{Corollary:} $a|b$ iff the remainder when $b$ is divided by $a$ is 0.

\textbf{Proof:} Let $b=aq+r$ by the division algorithm. If $r=0$, then $b=aq$ and so $a|b$. Conversely, if $a|b$, then $b=ak$ for some integer $k$. By the uniqueness part of the division algorithm, we must have $q=k$ and $r=0$. \qed

\subsubsection{2.6 - The GCD}

\underline{Fact:} If you have a nonzero integer $a$, then it has finite divisors by the well-ordering principle. (eg there are only finitely many positive integers less than or equal to $|a|$, and thus finite candidates for divisors). 

\textbf{Definition (GCD):} Let $a,b$ be integers, not both zero. The greatest common divisor of $a$ and $b$, denoted $\gcd(a,b)$, is the largest positive integer that divides both $a$ and $b$.

\textbf{Example:} \begin{itemize}
    \item $\gcd(12,15)=3$
    \item $\gcd(0,5)=5$
    \item $\gcd(0,0)$ is not defined.
\end{itemize}

\textbf{Definition (Relatively Prime):} Two integers are relatively prime if their gcd is 1.

\textbf{Example:} \begin{itemize}
    \item $\gcd(8,15)=1$ so 8 and 15 are relatively prime.
    \item $\gcd(9,28)=1$ so 9 and 28 are relatively prime.
    \item $\gcd(0,1)=1$ so 0 and 1 are relatively prime.
\end{itemize}

\textbf{Corollary:} $gcd(p,n)=1\iff p \nmid n$.

\textbf{Proof:} If $\gcd(p,n)=1$, then the only positive divisor of both $p$ and $n$ is 1. Thus, $p$ does not divide $n$. Conversely, if $p \nmid n$, then the only positive divisor of both $p$ and $n$ is 1, so $\gcd(p,n)=1$. \qed

\section{Lecture 4}

Last time, we defined the GCD. 

\textbf{Proposition (2.10):} Let $a,b\in \zz$ and $d=\gcd(a,b)$. Then, \begin{align*}
    \gcd(\frac{a}{d},\frac{b}{d})=1
\end{align*}

\textbf{Proof:} Recall that the gcd is defined to be at least 1, always. We will prove that \begin{align*}
    \gcd(\frac{a}{d},\frac{b}{d})\leq 1
\end{align*}

To do this, we will show that any common divisor of $\frac{a}{d}$ and $\frac{b}{d}$ is $\leq 1$. 

Suppose that for some positive integer $c$, $c|\frac{a}{d}$ and $c|\frac{b}{d}$. By def of divisibility we thus have that $\frac{a}{d}=c\cdot k$, and that $\frac{b}{d}=c\cdot l$ for $k,l\in \zz$.

Then, $a=c\cdot d \cdot k$, and $b=c\cdot d\cdot l$. This shows that $cd$ is a common divisor of $a$ and $b$. However, $d$ is the gcd of $a,b$, so it must be that $cd\leq d$. Under $\mathbb{Q}$ field properties there must $\exists!$ inverse to $d$, $\frac{1}{d}$ and by field closure we can multiply, so $c\leq 1$. \qed 

\subsection{2.7 - The Euclidean Algorithm}

How can we compute gcds?

One method is to factor each one into primes, take the overlapping subset, and find their product. This is dumb for large integers because factoring large integers is not computationally feasible (for traditional computers)

\textbf{Lemma:} Suppose $a=bq+r$. The set of common divisors of $a$ and $b$ is the same as the set of common divisors of $b$ and $r$. It thus follows they have the same gcd. 

\textbf{Proof:} Suppose $a=bq+r$. Then, $a$ is a linear combination of $b,r$. Then, any common divisor of $b,r$ must also divide $a$ by a former proposition. Thus, any common divisor of $b,r$ is by definition a divisor of $b$, but also a divisor of $a$. Thus, they are also common divisors of $a,b$. 

Similarly, we can rewrite the equation to say that $r=a-bq$, and thus similarly all common divisors of $a,b$ must be common divisors of $r,b$. So we are done. \qed 

\underline{Note:} Here we can see the idea for the euclidean algorithm presented. We can reduce the gcd problem from $a,b$ to be $r,b$, where $r$ is strictly lesser than $a$. In fact, it's $\leq b$ by definition of division algorithm. Then, we can do this to $b$, then $r$ again, etc until the numbers are very small. 

\underline{Example:} Let's use this idea to compute $\gcd(315,220)$. Then, \begin{align*}
    315=1\cdot 220 + 95
    220=2\cdot 95 + 30
    95 =3\cdot 30 + 5
    30 =6\cdot 5 + 0
\end{align*}
and so applying the lemma, we get that $\gcd(315,220)=\gcd(220,95)=\gcd(95,30)=\gcd(30,5)=\gcd(5,0)=5$.

or the last nonzero remainder, is the gcd. 

\textbf{Theorem (The Euclidean Algorithm):} Let $a,b\geq 0\in \zz,b\neq 0$. Then, $a=bq_1+r_1$ with $0\leq r_1\leq b$. Then, recursively apply the theorem with parameters $b,r$. The base case applies when the remainder is 0, in which case the divisor is the gcd. 

In python, 
\begin{lstlisting}[language=Python]
def gcd(a,b):
    if b==0:
        return a
    else:
        return gcd(b, a%b)
\end{lstlisting}

\textbf{Proof:} This is just repeated application of the lemma. \qed

\subsubsection{2.7.1 - The Extended Euclidean Algorithm}

\textbf{Theorem (2.12):} $\exists x,y\in \zz$ st for $a,b\in \zz$ not both zero, $ax+by=\gcd(a,b)$.

\underline{Note:} They are not unique. 

\section{Lecture 5}

The idea is that we can solve for the remainder. For instance, take the example:

\underline{Ex:} $\gcd(1239,735)$ results in \begin{align*}
    1239 &= 1\cdot 735 + 504 \\
    735 &= 1\cdot 504 + 231 \\
    504 &= 2\cdot 231 + 42 \\
    231 &= 5\cdot 42 + 21 \\
    42 &= 2\cdot 21 + 0
\end{align*}

But this also gives \begin{align*}
    504 &= 1239 - 1\cdot 735 \\
    231 &= 735 - 1\cdot 504 \\
    42 &= 504 - 2\cdot 231 \\
    21 &= 231 - 5\cdot 42
\end{align*}

We can substitute these all the way up to get \begin{align*}
    21 &= 231 - 5\cdot 42 \\
    &= 231 - 5\cdot (504 - 2\cdot 231) \\
    &= 11\cdot 231 - 5\cdot 504 \\
    &= 11\cdot (735 - 1\cdot 504) - 5\cdot 504 \\
    &= 11\cdot 735 - 16\cdot 504 \\
    &= 11\cdot 735 - 16\cdot (1239 - 1\cdot 735) \\
    &= 27\cdot 735 - 16\cdot 1239
\end{align*}

This gives us $x=-16$ and $y=27$ such that $1239\cdot (-16) + 735\cdot 27 = 21$.

\textbf{Theorem (The Extended Euclidean Algorithm):} Let $a,b\in \zz$ not both zero. Then, we can find $x,y\in \zz$ such that $ax+by=\gcd(a,b)$. Do so by: \begin{itemize}
    \item Apply the Euclidean algorithm to find the gcd and all the remainders.
    \item Rewrite each remainder as a linear combination of $a$ and $b$ by substituting the previous remainders.
    \item Continue substituting until you express the gcd as a linear combination of $a$ and $b$.
\end{itemize}

In python, 
\begin{lstlisting}[language=Python]
def extended_gcd(a,b):
    if b==0:
        return (a,1,0)
    else:
        d,x1,y1 = extended_gcd(b, a%b)
        x = y1
        y = x1 - (a//b)*y1
        return (d,x,y)
\end{lstlisting}

\textbf{Proof:} This is just the algorithm described above. The correctness follows from the Euclidean algorithm and properties of linear combinations. \qed

\underline{Example:} Find $x,y$ such that $19x+7y=1$. 

We apply the extended euclidean algorithm: \begin{align*}
    19 &= 2\cdot 7 + 5 \\
    7 &= 1\cdot 5 + 2 \\
    5 &= 2\cdot 2 + 1 \\
    2 &= 2\cdot 1 + 0
\end{align*}

Then, in reverse, \begin{align*}
    1 &= 5 - 2\cdot 2 \\
      &= 5 - 2\cdot (7 - 1\cdot 5) \\
      &= 3\cdot 5 - 2\cdot 7 \\
      &= 3\cdot (19 - 2\cdot 7) - 2\cdot 7 \\
      &= 3\cdot 19 - 8\cdot 7
\end{align*}

Some applications:

\textbf{Proposition (2.16):} Let $a,b,c\in \zz$ st $a\neq 0, \gcd(a,b)=1$. Then if $a|bc$, then $a|c$.

The idea is that if $a,b$ are coprime, then $a$ has no common factors with $b$. Thus, if $a$ divides $bc$, it must be that $a$ divides $c$.

\textbf{Proof:} Since $\gcd(a,b)=1$, there exist $x,y\in \zz$ such that $ax+by=1$. Multiplying both sides by $c$ gives $acx + bcy = c$. Since $a|acx$ and $a|bcy$ (since $a|bc$), it follows that $a|c$. \qed

\textbf{Corollary:} Let $p\in \zz$ prime, and $a,b\in \zz$. Then if $p|ab$, then $p|a$ or $p|b$.

\textbf{Proof:} If $p|a$, we are done. Otherwise, $\gcd(p,a)=1$ since $p$ is prime and does not divide $a$. Thus, by the previous proposition, $p|b$. \qed

\textbf{Corollary (2.14):} Let $a,b,d\in \zz$ st $d=\gcd(a,b)$. For $e\in \zz$ if $e|a, e|b$, then $e|d$.

\textbf{Proof:} Since $d$ is a linear combination of $a$ and $b$, we have that $d = ax + by$ for some integers $x,y$. Since $e|a$ and $e|b$, it follows that $e|ax$ and $e|by$. Thus, $e|d$. \qed

\subsection{2.9 - Fermat and Mersenne Primes}

We know that there are infinitely many primes by Euclid's theorem. However, are there infinitely many primes of certain forms? And can we use this to generate primes? Useful for cryptography, etc. 

\textbf{Definition (Mersenne Number):} The $n$th Mersenne number is defined as $M_n = 2^n - 1$ for $n\geq 1$.

\textbf{Proposition:} If $a|n$, then $M_a|M_n$. If $n$ is composite, then $M_n$ is composite.

\textbf{Proof:} Suppose $n=ab$ for integers $a,b>1$. Then, \begin{align*}
    M_n &= 2^{ab} - 1 \\
        &= (2^a)^b - 1^b \\
        &= (2^a - 1)\left( (2^a)^{b-1} + (2^a)^{b-2} + \cdots + 1 \right)
\end{align*}
Since $a,b>1$, it follows that $2^a - 1 > 1$ and the second factor is also greater than 1. Thus, $M_n$ is composite. \qed

\underline{Example:} $M_10 = 2^{10} - 1 = 1023$ is composite. We know that $2|10$ so $M_2|M_{10}$. In fact, $M_2=3$ and $1023=3\cdot 341$. Also, $5|10$ so $M_5|M_{10}$. In fact, $M_5=31$ and $1023=31\cdot 33$.

\underline{Note:} If $p$ is prime, is $M_p$ prime? For some, yes, i.e. 13, 17, 19. However $M_{11}$ isn't prime for instance. 

\textbf{Definition (Fermat Number):} The $n$th Fermat number is defined as $F_n = 2^{2^n} + 1$ for $n\geq 0$.

\underline{Note:} He initially looked at $2^m+1$ but found that unless $m$ is a power of 2, $2^m+1$ is composite. For instance, $2^6+1=65$ is composite.

\textbf{Proposition:} For $2^m+1$, if $m$ is not a power of 2, then $2^m+1$ is composite. If $m=2^n$, then $2^m+1$ is prime for $n=0,1,2,3$ but composite for $n=4$.

\textbf{Proof:} Suppose $m$ is not a power of 2. Recall that if $k$ is odd, then $x^k+1$ is divisible by $x+1$. Since $m$ is not a power of 2, then we can write $m=2^n\cdot k$ for some odd integer $k>1$. Then, \begin{align*}
    2^m + 1 &= 2^{2^n\cdot k} + 1 \\
            &= (2^{2^n})^k + 1^k \\
            &= (2^{2^n} + 1)\left( (2^{2^n})^{k-1} - (2^{2^n})^{k-2} + \cdots - 1 \right)
\end{align*}
Since $k>1$, it follows that $2^{2^n} + 1 > 1$ and the second factor is also greater than 1. Thus, $2^m + 1$ is composite. \qed

\section{Lecture 6}

\textbf{Proposition:} If $M_p=2^p-1$ is prime, then $p$ is prime. Such an $M_p$ is called a Mersenne prime.

\underline{Note:} There are only 52 known Mersenne primes! And it's not known if there are infinitely many.

It turns out that $F_5, F_6, F_7$ are all composite. In fact, only $F_0$ to $F_4$ are known to be prime. It is not known if there are infinitely many Fermat primes.

\section{Lecture 7}

Chapter 3 now, Linear Diophantine Equations.

\subsection{3.1 - Linear Diophantine Equations}

Take $ax+by=c$ for $a,b,c\in \zz$. We are interested in integer pairs $(x,y)$ that satisfy this linear diophantine equation.

The goal will be to describe \underline{all} such solutions $(x,y)$

\underline{Example:} Find integers $x,y$ such that $19x+7y=1$.

By the extended euclidean algorithm, \begin{align*}
    19 &= 2\cdot 7 + 5 \\
    7 &= 1\cdot 5 + 2 \\
    5 &= 2\cdot 2 + 1 \\
    2 &= 2\cdot 1 + 0
\end{align*}

Then, in reverse, \begin{align*}
    1 &= 5 - 2\cdot 2 \\
      &= 5 - 2\cdot (7 - 1\cdot 5) \\
      &= 3\cdot 5 - 2\cdot 7 \\
      &= 3\cdot (19 - 2\cdot 7) - 2\cdot 7 \\
      &= 3\cdot 19 - 8\cdot 7
\end{align*}

Then, one solution is $3, -8$. There are infinitely many solutions, since we can add $7k$ to $x$ and subtract $19k$ from $y$ for any integer $k$ to get another solution. So, the set of all solutions is $\{ (3+7k, -8-19k) : k\in \zz \}$.

\underline{Example:} Find integers $x,y$ such that $19x+7y=6$. 

We do know that \begin{align*}
    1&= 3\cdot 19 - 8\cdot 7 \\
    6&= 18\cdot 19 - 48\cdot 7 \\
\end{align*}

Then, we can similarly add $7k$ to $x$ and subtract $19k$ from $y$ for any integer $k$ to get another solution. So, the set of all solutions is $\{ (18+7k, -48-19k) : k\in \zz \}$.

\underline{Example:} Find integers $x,y$ such that $6x+15y=10$. 

There are no integer solutions since $\gcd(6,15)=3$ does not divide 10. Generally if $\gcd(a,b)$ does not divide $c$, then there are no integer solutions to $ax+by=c$.

\textbf{Proof:} Let $d=\gcd(a,b)$. Then $d|a, d|b$ and $d|ax+by$ by bilinearity of divisibility. However, $d\nmid c$ by assumption, so there are no integer solutions to $ax+by=c$. \qed

\textbf{Theorem (3.1):} Assume $a,b,c\in \zz$, and at least one of $a,b\neq 0$, and let $d=\gcd(a,b)$. Then, \begin{enumerate}[(1)]
    \item The equation $ax+by=c$ has integer solutions iff $d|c$.
    \item Additionally, if it has an integer solution, it has infinitely many integer solutions. In particular, if $(x_0,y_0)$ is one integer solution, then the set of all integer solutions is given by \begin{align*}
        \{ (x_0 + \frac{b}{d}k, y_0 - \frac{a}{d}k) : k\in \zz \}
    \end{align*}
\end{enumerate}

\textbf{Proof:} \begin{enumerate}[(1)]
    \item $\Rightarrow$: Let $ax+by=c$ have integer solutions. Then, since $d|a,d|b$, we have that $d|ax+by$ by bilinearity of divisibility. Thus, $d|c$.
    
    $\Leftarrow$ Suppose $d|c$. THen, $c=d\cdot e$ for some integer $e$. Since $d$ is a linear combination of $a$ and $b$, we have that $d = ax_0 + by_0$ for some integers $x_0,y_0$. Thus, \begin{align*}
        c &= d\cdot e \\
          &= (ax_0 + by_0)e \\
          &= a(ex_0) + b(ey_0)
    \end{align*}
    So, $(ex_0, ey_0)$ is an integer solution to $ax+by=c$.
    \item (This is my own proof) First, $\frac{b}{d}, \frac{a}{d}\in \zz$ by proposition 2.10. Now, let $(x_0,y_0)$ be one integer solution to $ax+by=c$. Then, for any integer $k$, we have \begin{align*}
        a(x_0 + \frac{b}{d}k) + b(y_0 - \frac{a}{d}k) &= ax_0 + by_0 + ab\frac{k}{d} - ab\frac{k}{d} \\
        &= c
    \end{align*}
    Thus, $(x_0 + \frac{b}{d}k, y_0 - \frac{a}{d}k)$ is an integer solution for any integer $k$.
\end{enumerate}\qed

\textbf{Definition (Homogeneous LDE):} A linear diophantine equation of the form $ax+by=0$ is called a homogeneous linear diophantine equation.

\textbf{Lemma:} Suppose $a,b$ are not both 0. Then \begin{enumerate}[(1)]
    \item If $\gcd(a,b)=1$, every solution to $ax+by=0$ is of the form $(bk, -ak)$ for some integer $k$.
    \item In general, if $d=\gcd(a,b)$, every solution to $ax+by=0$ is of the form $(\frac{b}{d}k, -\frac{a}{d}k)$ for some integer $k$.
\end{enumerate}

\textbf{Proof:} \begin{enumerate}[(1)]
    \item Let $a,b$ not both 0. Suppose $\gcd(a,b)=1$. Let $(x,y)$ be a solution to $ax+by=0$. Then, $ax=-by$ and so $a|by$. Since $\gcd(a,b)=1$, it follows that $a|y$. Thus, $y=ak$ for some integer $k$. Substituting back gives $ax=-bak$, so $x=-bk$. Thus, every solution is of the form $(bk, -ak)$ for some integer $k$.
    \item Let $a,b$ not both 0. Let $d=\gcd(a,b)$. Let $(x,y)$ be a solution to $ax+by=0$. Then, $ax=-by$ and so $a|by$. Since $\gcd(a,b)=d$, it follows that $\frac{a}{d}|\frac{b}{d}y$. Since $\gcd(\frac{a}{d}, \frac{b}{d})=1$ by proposition 2.10, it follows that $\frac{a}{d}|y$. Thus, $y=\frac{a}{d}k$ for some integer $k$. Substituting back gives $ax=-b\frac{a}{d}k$, so $x=-\frac{b}{d}k$. Thus, every solution is of the form $(\frac{b}{d}k, -\frac{a}{d}k)$ for some integer $k$.
\end{enumerate}

\section{Lecture 8}

\textbf{Theorem (3.1):} Let $d=\gcd(a,b)$. Then \begin{enumerate}
    \item $ax+by=c$ has solutions iff $d|c$. 
    \item If $(x_0,y_0)$ is one solution to $ax+by=c$, then the solution set is given by \begin{align*}
        x=x_0+\frac{b}{d}t, y=y_0-\frac{a}{d}t
    \end{align*}where $t$ is an integer
\end{enumerate}

We proved this last time, along with a related lemma. I will now write Prof's proof of part 2.

\textbf{Proof:} (of part 2) Suppose $(x_0,y_0)$ is some solution for $ax+by=c$. Then, we should show that the above format gives solutions, and additionally that they are all of the solutions. So \begin{itemize}
    \item Verify it is a solution: \begin{align*}
        a(x_0+\frac{b}{d}t)+b(y_0+\frac{a}{d}t)&=ax_0+\frac{ab}{d}t+by_0-\frac{ab}{d}t \\
        &= ax_0+by_0 
    \end{align*}
    which is a solution by assumption. 
    \item Verify it is all the solutions: Let $(x,y)$ be any solution to $ax+by=c$. Then, \begin{align*}
        a(x-x_0)+b(y-y_0) &= ax+by - (ax_0+by_0) \\
        &= c-c \\
        &= 0
    \end{align*}
    Thus, $(x-x_0,y-y_0)$ is a solution to the homogeneous linear diophantine equation $ax+by=0$. By the previous lemma, there exists an integer $t$ such that \begin{align*}
        (x-x_0,y-y_0) &= (\frac{b}{d}t, -\frac{a}{d}t)
    \end{align*}
    Thus, \begin{align*}
        (x,y) &= (x_0+\frac{b}{d}t, y_0-\frac{a}{d}t)
    \end{align*}
    which is of the desired form.
\end{itemize}\qed 

\underline{Note:} Thus, to solve a linear diophantine equation \begin{align*}
    ax+by=c
\end{align*}, first ensure that $d=\gcd(a,b)$ divides $c$. If not, there are no solutions. If it does, then find one solution $(x_0,y_0)$ using the extended euclidean algorithm. Then, the set of all solutions is given by \begin{align*}
    \{ (x_0 + \frac{b}{d}t, y_0 - \frac{a}{d}t) : t\in \zz \}
\end{align*}

\underline{Example:} Solve $12x+45y=30$. Then, $\gcd(12,45)=3, 3|30$ so there are solutions \begin{align*}
    45 &= 12(3)+9\\
    12 &= 9(1)+3 \\
    9 &= 3(3)+0
\end{align*}
going back up we have \begin{align*}
    3 &= 12 - 9(1) \\
      &= 12 - (45 - 12(3)) \\
      &= 4\cdot 12 - 45
\end{align*}
we can multiply by 10 to get $30 = 40\cdot 12 - 10\cdot 45$. Thus, one solution is $(40,-10)$. The set of all solutions is given by \begin{align*}
    \{ (40 + 15t, -10 - 4t) : t\in \zz \}
\end{align*}

\subsection{3.2 - Postage Stamp Problem}

\textbf{Problem:} You have an unlimited supply of 3 cent and 5 cent stamps. What amounts of postage can you make? Generalize this to $a$ cent and $b$ cent stamps, frequently where $a,b$ are coprime.

The answer is obviously intuitively \begin{align*}
    ax+by&=c
\end{align*} for some nonnegative integers $x,y$. We want to find the largest $c$ such that there are no nonnegative integer solutions to this equation. Also, what are all of the values of $c$ such that there are no nonnegative integer solutions to this equation?

\begin{aside}
Professor gives an example for postage stamps with values of 3 and 5 here, which is a trivially solved problem. The idea is that if you can find 3 consecutive values of $c$ that can be achieved, then all values of $c$ greater than the largest of those 3 must also be achievable. This is because you can add 3 to any of those 3 values to get the next value, and so on. Thus, you just need to find the largest $c$ such that there are no 3 consecutive values of $c$ that can be achieved.

In general, we can intuit from this that for the smallest of $a,b$, if we can find $a$ consecutive values of $c$ that can be achieved, then all values of $c$ greater than the largest of those $a$ must also be achievable. Thus, you just need to find the largest $c$ such that there are no $a$ consecutive values of $c$ that can be achieved.
\end{aside}

\underline{Example:} Suppose in some fucked up version of football the only ways to score are with 3 or 7 points. What scores can be achieved?

Remember you just need 3 in a row that can be achieved and everything after is doable. 

\begin{align*}
    \begin{tabular}{c|c}
        Score & Achievable? \\
        \hline
        0 & Yes \\
        1 & No \\
        2 & No \\
        3 & Yes \\
        4 & No \\
        5 & No \\
        6 & Yes \\
        7 & Yes \\
        8 & No \\
        9 & Yes \\
        10 & Yes \\
        11 & No \\
        12 & Yes \\
        13 & Yes \\
        14 & Yes \\
    \end{tabular}
\end{align*}and so on. We can see that the largest score that cannot be achieved is 11, and all scores greater than 11 can be achieved.

\textbf{Proposition (3.3):} Suppose $a>1, b>1$ are coprime. Then there are no nonnegative integers $x,y$ such that $ax+by=ab-a-b$. 

\underline{Note:} The general idea behind this proof is that when you plug in any arbitrary integer solution to $ax+by=ab-a-b$, you get that $ax_0+by_0=1$. Since $a,b>1$, this forces $x_0,y_0$ to be nonnegative and sum to at most 1, none of which can satisfy the equation.

\textbf{Proof:} Since $a,b$ are coprime, there exist integers by a previous theorem that $ax_0+by_0=1$. 

We know that $(ab-a-b)x_0,(ab-a-b)y_0$ is a solution to $ax+by=ab-a-b$. Since $a,b>1, ab-a-b>0$, and so for both to be positive we must have $x_0,y_0\geq 0$. 

Thus, $ax_0+by_0\geq 0$. However, since $x_0,y_0$ are integers, they must sum to at most 1, otherwise $ax_0+by_0\geq 2a+2b>1$. Thus, $0\leq x_0+y_0\leq 1$. By integer properties, this means that either $x_0=0,y_0=0$ or one of them is 1 and the other is 0. 

Then, if $x_0=0,y_0=0$, then $ax_0+by_0=0\neq 1$. If $x_0=1,y_0=0$, then $ax_0+by_0=a>1$. If $x_0=0,y_0=1$, then $ax_0+by_0=b>1$. But we need it to be identically 1 by assumption, a contradiction.\qed

\begin{aside}
The professor gives a proof that requires memorizing some answer for $x_0,y_0$, which is why I derived my own. 

\textbf{Proof:} (by the professor) Consider the Diophantine Equation \begin{align*}
    ax+by=ab-a-b
\end{align*}Notice that $x_0=-1, y_0=a-1$ is a solution, as \begin{align*}
    a(-1) + b(a-1) &= -a + ab - b \\
                   &= ab - a - b
\end{align*}
Thus, the general solution of the Diophantine Equation is $x=-1+bt,y=a-1-at$ for some integer $t$. 

WLOG, suppose $(x,y)$ is a solution such that $x\geq 0$. Since $x\geq 0$, we have that $-1+bt\geq 0$. This implies that $t>0$, an integer, so $t\geq 1$. Then, since $y=a-1-at$, we have that $y\leq a-1-a=-1<0$. Thus, if $x\geq 0$, then $y<0$. Similarly, if $y\geq 0$, then $x<0$. Thus, there are no nonnegative integer solutions to the equation. \qed
\end{aside}

\section{Lecture 9}

\underline{Recall:} The postage stamp problem asks, given $a,b$ coprime positive integers, what is the largest integer $c$ such that there are no nonnegative integer solutions to $ax+by=c$?

\textbf{Proposition (3.4):} Let $a,b$ be coprime positive integers. If $n\geq ab-a-b+1$ then there are nonnegative integers $x,y,$ such that $ax+by=n$.

\textbf{Proof:} Let $(x_0,y_0)$ be an integer solution to $ax+by=n$, which exists because $\gcd(a,b)=1$ divides $n$. Then, the set of all integer solutions is given by \begin{align*}
    \{ (x_0 + bk, y_0 - ak) : k\in \zz \}
\end{align*}

Divide $y_0$ by $a$ to get $y_0=qa+r$ for some integer $q$ and $0\leq r<a$ by the division algorithm. Consider the solution $(x_1,y_1)$ corresponding to $k=q$, which is given by \begin{align*}
    (x_1,y_1) &= (x_0 + bq, y_0 - ak) \\
             &= (x_0 + bq, r)
\end{align*}Thus, $0\leq y_1<a$. We need to show that $x_1\geq 0$. Note that we have that $n\geq ab-a-b-1$, and also that $y_1\leq a-1$ because $y_1=r<a$. This implies $-y_1\geq -(a-1)$. Since $ax_1+by_1=n$, we have \begin{align*}
    x_1&=\frac{n-by_1}{a} \\
       &\geq \frac{ab-a-b+1 - b(a-1)}{a} \\
       &= \frac{ab-a-b+1 - ab + b}{a} \\
       &= \frac{1-a}{a} \\
       &= -1 + \frac{1}{a}
\end{align*}But because $x_1$ is an integer, it must be that $x_1\geq 0$. Thus, there are nonnegative integer solutions to $ax+by=n$. \qed

\textbf{Corollary:} From the above two propositions, the largest integer $c$ such that there are no nonnegative integer solutions to $ax+by=c$ is $ab-a-b$. That is, for all $n\geq ab-a-b+1$, there are nonnegative integer solutions to $ax+by=n$, and for $n=ab-a-b$, there are no nonnegative integer solutions to $ax+by=n$. 

\textbf{Proof:} Compose the previous 2 proofs.

\underline{Example:} The 2026 Postage Stamp Problem uses values of: \begin{align*}
    \begin{tabular}{c|c}
        \text{Stamp} & \text{Stamp Value}\\
        \hline
        \text{Regular} & 78\\
        \text{Postcard} & 61
    \end{tabular}
\end{align*}

First, we must have that they are coprime. Since $78=2\cdot 3\cdot 13$ and $61$ is prime, they are coprime.

Then, under this problem, the largest amount of postage that cannot be made is $78\cdot 61 - 78 - 61 = 4619$. For all amounts of postage greater than 4619, there are nonnegative integer solutions to $78x+61y=n$.

\underline{Example:} Find the smallest for stamp values of $6, 8$. These two are not coprime, so we will divide the equation by their gcd (2). Then, we find the largest amount of postage that cannot be made with stamps of values $3,4$, which is $3\cdot 4 - 3 - 4 = 5$. Then, we multiply this by the gcd (2) to get $10$ as the largest amount of postage that cannot be made with stamps of values $6,8$. For all amounts of postage greater than 10, there are nonnegative integer solutions to $6x+8y=n$.

\subsection{QUIZ 1 CUTOFF HERE}

\subsection{Appendix A.3 - Well-Ordering Principle and Mathematical Induction}

\textbf{Theorem (Well-Ordering Principle):} Every nonempty subset of $\mathbb{N}$ (the positive integers) has a least element.

\underline{Note:} This only applies to the naturals and not the integers, since the integers can be negative and thus have no least element. Most notably the proof of the well-ordering principle uses the principle of succession, which is a property of the naturals specifically due to Peano succession axioms. You can also prove this utilizing the completeness of the reals, but once again this requires the existence of integers specifically in a half open interval.

Specifically it requires the completeness property, showing that all bottom-bounded subsets of the reals have a least upper bound, and because of the transitivity of $\subseteq$, for $A\subseteq \mathbb{N}$, $A$ has an infinum called $a$, and then $(n-1,n]$ must contain an element of $A$ for all $n\geq a$, so there must be some least element of $A$.

\underline{Note:} We will use this to give a non-constructive proof of the following theorem:

\textbf{Theorem (2.12):} Let $a,b$ be integers, not both 0 and let $d=\gcd(a,b)$. Then, there exist integers $x,y$ such that $ax+by=d$.

\underline{Note:} Last time, we proved this using the Extended Euclidean Algorithm, which is a constructive proof. This time, we will give a non-constructive proof using the Well-Ordering Principle.

\textbf{Proof:} Let $S=\{ ax+by : x,y\in \zz, ax+by>0 \}$ be the set of all positive linear combinations of $a$ and $b$. Since $a,b$ are not both 0, there exists some positive linear combination of $a$ and $b$, so $S$ is nonempty. By the Well-Ordering Principle, $S$ has a least element, say $d$. We will show that $d=\gcd(a,b)$.

First, we show that $d|a$ and $d|b$. By the division algorithm, we can write $a=dq+r$ for some integers $q,r$ with $0\leq r<d$. Then, \begin{align*}
    r &= a - dq \\
      &= a - q(ax_0 + by_0) \\
      &= a(1-qx_0) + b(-qy_0)
\end{align*}where $x_0,y_0$ are integers such that $d=ax_0+by_0$. Thus, $r$ is a linear combination of $a$ and $b$. If $r>0$, then $r\in S$ and $r<d$, contradicting the minimality of $d$. Thus, $r=0$ and so $d|a$. Similarly, we can show that $d|b$.

\begin{aside}
Professor starts a contradiction by stating suppose $r\neq 0$ and then doing the above, instead of taking cases. 
\end{aside}

Next, we show that if $e$ is any common divisor of $a$ and $b$, then $e|d$. Since $e|a$ and $e|b$, it follows that $e|ax+by$ for any integers $x,y$. In particular, since $d=ax_0+by_0$, it follows that $e|d$. Since $d>0$, $e\leq d$. Thus, $d$ is the greatest common divisor of $a$ and $b$. \qed

\underline{Note:} This argument shows that $\gcd(a,b)$ is the least positive linear combination of $a$ and $b$.

\underline{Remark:} Let $d=\gcd(a,b)$. 

\section{Lecture 10}

\underline{Induction:} Let $P(n)$ be a statement about a positive integer $n$. For instance, $n$ is a product of primes. Then, if $P(1)$ is true, and for all $n\geq 1$, if $P(n)$ is true, then $P(n+1)$ is true, then $P(n)$ is true for all positive integers $n$.

The induction scheme expressed in the above is called weak induction. There is also strong induction, which states that if $P(1)$ is true, and for all $n\geq 1$, if $P(1), P(2), \ldots, P(n)$ are all true, then $P(n+1)$ is true, then $P(n)$ is true for all positive integers $n$.

Note that this means that for strong induction, we only have to prove $n=1$, but we can \textit{assume} that $P(1), P(2), \ldots, P(n)$ are all true to prove $P(n+1)$. For weak induction, we have to prove $n=1$ and then for each $n\geq 1$, we have to prove that if $P(n)$ is true, then $P(n+1)$ is true. Thus, strong induction is often easier to use than weak induction. 

\begin{aside}
I think we are allowed to do so because of the well-ordering principle, which states that every nonempty subset of the positive integers has a least element. If there is some $n$ such that $P(n)$ is false, then there is a least such $n$. Since $P(1)$ is true, $n>1$. Since $n$ is the least such integer, $P(1), P(2), \ldots, P(n-1)$ are all true. Thus, by the strong induction hypothesis, $P(n)$ is true, a contradiction. Thus, there are no such integers $n$ such that $P(n)$ is false, and so $P(n)$ is true for all positive integers $n$.
\end{aside}

Additionally, we can express them from a logic perspective:\begin{itemize}
    \item $P(1) \wedge \forall n\geq 1, P(n)\rightarrow P(n+1) \rightarrow \forall n\geq 1, P(n)$
    \item $P(1) \wedge \forall n\geq 1, (P(1)\wedge P(2)\wedge \cdots \wedge P(n))\rightarrow P(n+1) \rightarrow \forall n\geq 1, P(n)$
\end{itemize}

\underline{Example:} Let's prove that $1+3+5+\dots+(2n-1)=n^2$ for all $n\geq 1$. That is, \begin{align*}
    \sum_{k=1}^n (2k-1) = n^2
\end{align*}

\textbf{Proof:} We will use weak induction. For the base case, $n=1$, we have $1=1^2$, so the base case holds. Now, suppose that for some $n\geq 1$, we have $\sum_{k=1}^n (2k-1) = n^2$. Then, \begin{align*}
    \sum_{k=1}^{n+1} (2k-1) &= \sum_{k=1}^n (2k-1) + (2(n+1)-1) \\
                             &= n^2 + 2n + 1 \\
                             &= (n+1)^2
\end{align*}Thus, by the principle of weak induction, we have that $\sum_{k=1}^n (2k-1) = n^2$ for all $n\geq 1$. \

\underline{Example:} Prove that $n!\geq n\cdot 2^n$ for all $n\geq 6$. 

\textbf{Proof:} By induction, \begin{itemize}
    \item \underline{Base:} $n=6$, $6!=720\geq 6\cdot 2^6=384$, so the base case holds.
    \item \underline{Induction Step:} Suppose that for some $n\geq 6$, we have $n!\geq n\cdot 2^n$. Then, \begin{align*}
        (n+1)! &= (n+1)n! \\
               &\geq (n+1)n\cdot 2^n \\
               &\geq (n+1)\cdot 2^{n+1}
    \end{align*}Thus, by the principle of weak induction, we have that $n!\geq n\cdot 2^n$ for all $n\geq 6$. \qed
\end{itemize}

\underline{Example:} Prove that $3|(n^3-n)$ for all $n\geq 0$

\textbf{Proof:} By induction, \begin{itemize}
    \item \underline{Base:} $n=0$, $3|0^3-0=0$, so the base case holds.
    \item \underline{Induction Step:} Suppose that for some $n\geq 0$, we have $3|(n^3-n)$. Then, \begin{align*}
        (n+1)^3-(n+1) &= n^3 + 3n^2 + 3n + 1 - n - 1 \\
                      &= n^3 - n + 3n^2 + 3n \\
                      &= (n^3 - n) + 3(n^2 + n)
    \end{align*}Since $3|(n^3-n)$ by the induction hypothesis, and $3|(3(n^2+n))$ by bilinearity of divisibility, it follows that $3|((n^3-n) + 3(n^2+n))$. Thus, by the principle of weak induction, we have that $3|(n^3-n)$ for all $n\geq 0$. \qed
\end{itemize}

\underline{Note:} Replace 3 with any prime $p$ to get that $p|(n^p-n)$ for all $n\geq 0$. This is a special case of Fermat's Little Theorem, which states that if $p$ is prime and $a$ is an integer not divisible by $p$, then $a^{p-1}\equiv 1 \mod p$.

\textbf{Lemma (4.3):} Every integer $n\geq 2$ can be written as a product of primes \begin{align*}
    n=p_1p_2\cdots p_k
\end{align*}

\textbf{Proof:} We will use strong induction on $n$.\begin{itemize}
    \item \underline{Base:} For $n=2$, $2=2$ is a product of primes, so the base case holds.
    \item \underline{Inductive Step:} Now assume every integer $k$ with $2\leq k\leq n$ can be written as a product of primes. We want to show that $n+1$ can be written as a product of primes. If $n+1$ is prime, then we are done. If $n+1$ is composite, then there exist integers $a,b$ such that $2\leq a,b\leq n$ and $n+1=ab$. By the induction hypothesis, we can write $a=p_1p_2\cdots p_k$ and $b=q_1q_2\cdots q_m$ for some primes $p_i,q_j$. Thus, \begin{align*}
    n+1 &= ab \\
        &= (p_1p_2\cdots p_k)(q_1q_2\cdots q_m) \\
        &= p_1p_2\cdots p_k q_1q_2\cdots q_m
\end{align*}which is a product of primes. Thus, by the principle of strong induction, every integer $n\geq 2$ can be written as a product of primes. \qed
\end{itemize}

\section{Lecture 11}

Recall last time we did 3 main things; the well ordering principle, induction, and strong induction. 

All 3 of the above are equivalent to each other. This potentially means that you ould rewrite one proof using the others!

\subsection{4.1 - The Starting Point (of Unique Factorizations)}

\textbf{Theorem (4.1):} Let $p$ be a prime and $a,b$ be integers such that $p|ab$. Then, $p|a$ or $p|b$.

\textbf{Proof:} We proved this above in the euclidean algorithm process

\textbf{Corollary (4.2):} Let $p$ be a prime and $a_1,a_2,\ldots, a_k$ be integers such that $p|a_1a_2\cdots a_k$. Then, there exists some $i$ such that $p|a_i$.

\textbf{Proof:} We use induction on $k$.

\underline{Base:} If $k=2$ then this is just the previous theorem, so the base case holds. If $k=1$, then $p|a_1$ so the base case holds.

\underline{IS:} Now, suppose that for some $k\geq 2$, if $p|a_1a_2\cdots a_k$ then there exists some $i$ such that $p|a_i$. We want to show that if $p|a_1a_2\cdots a_{k+1}$ then there exists some $i$ such that $p|a_i$. If $p|a_{k+1}$, then we are done. If not, then since $p|a_1a_2\cdots a_{k+1}$ and $p\nmid a_{k+1}$, it follows that $p|(a_1a_2\cdots a_k)$ by bilinearity of divisibility. By the induction hypothesis, there exists some $i$ such that $p|a_i$. 
\qed

\subsection{4.2 - The Fundamental Theorem of Arithmetic}

\textbf{Theorem:} Every integer $n\geq 2$ can be written as a product of primes in a unique way, up to the order of the factors. That is, if \begin{align*}
    n &= p_1p_2\cdots p_k \\
      &= q_1q_2\cdots q_m
\end{align*}where $p_i,q_j$ are primes, then $k=m$ and there exists a permutation $\sigma$ of $\{1,2,\ldots, k\}$ such that $p_i=q_{\sigma(i)}$ for all $i$.

\textbf{Proof:} We already proved the existence of prime factorizations. 

\underline{Uniqueness:} First, observe that if $p$P is prime, then $p=p$ is a prime factorization. This is the only way to write $p$ as a product of primes, since if \begin{align*}
    p=p_1p_2\cdots p_k
\end{align*}is a prime factorization, then $p|p_1p_2\cdots p_k$, so there exists some $i$ such that $p|p_i$. Since $p_i$ is prime, it follows that $p=p_i$. Thus, the only prime factorization of a prime is itself.

Now, let's use strong induction on $n$. \begin{itemize}
    \item \underline{Base Case:} at $n=2$, $2$ is prime, so the only prime factorization of $2$ is $2$ itself, so the base case holds.
    \item \underline{Inductive Step:} Suppose $k$ is a number such that all $n$ such that $2\leq n\leq k$ have a unique prime factorization. Suppose $k+1$ has two prime factorizations \begin{align*}
        p_1p_2\cdots p_m &= q_1q_2\cdots q_n
    \end{align*}
    Then, $p_1|(q_1q_2\cdots q_n)$, so there exists some $i$ such that $p_1|q_i$. Since $q_i$ is prime, it follows that $p_1=q_i$. We can cancel $p_1$ and $q_i$ from both sides to get \begin{align*}
        p_2\cdots p_m &= q_1\cdots q_{i-1} q_{i+1}\cdots q_n
    \end{align*}
    By the induction hypothesis, the remaining factors must also be equal up to a permutation. Thus, the prime factorization of $k+1$ is unique.
\end{itemize}

\subsection{EXAM 1 CUTOFF HERE}

\section{Lecture 12}

Exam on Friday, the 27th. Covers up to Fundamental Theorem of Arithmetic.

HW due Wednesday

\underline{Note:} Often, we will write the prime factorization of an integer $n$ as \begin{align*}
    n&=p_1^{a_1}p_2^{a_2}\dots p_k^{a_k}
\end{align*}where all $p_i$ are unique, and $a_i$ are integers strictly greater than zero (eg $\geq 1$)

Sometimes, we'll write a prime factorizatin of an integer $n\geq 2$ as follows:\begin{align*}
    2^{n_2}3^{n_3}5^{n_5}\dots
\end{align*}which resembles an "infinite product" where each prime shows up. Then, each $n_i$ \textit{can} be zero. Additionally we must then have that only finitely many $n_p$ are nonzero. 

\underline{Example:} We know that $20=2^2\cdot 5$. This is equal to \begin{align*}
    2^2\cdot 3^0 \cdot 5^1 \cdot 7^0 \dots
\end{align*}

In this case $n_2=2, n_3=0, n_5=1, n_p=0$ for primes $p\geq 7$. 

\underline{Remark:} We can write $n=1$ this way too, with $n_p=0$ for all $p$. 

\textbf{Proposition (4.4):} Let $a,b$ be positive integers with prime factorizations \begin{align*}
    a=2^{a_2}3^{a_3}5^{a_5}\dots \quad b=2^{b_2}3^{b_3}5^{b_5}\dots
\end{align*}
Then, $a|b$ iff $a_p\leq b_p$ for every prime $p$. 

\textbf{Proof:} $\Rightarrow$ If $a|b$ then $b=ac$ for some integer $c\geq 1$. Write $c=2^{c_2}3^{c_3}5^{c_5}\dots$ Then, \begin{align*}
    b &= ac \\
      &= (2^{a_2}3^{a_3}5^{a_5}\dots)(2^{c_2}3^{c_3}5^{c_5}\dots) \\
      &= 2^{a_2+c_2}3^{a_3+c_3}5^{a_5+c_5}\dots
\end{align*}By the fundamental theorem of arithmetic, the prime factorization of $b$ is unique, so $b_p=a_p+c_p$ for every prime $p$. Since $c_p\geq 0$, it follows that $a_p\leq b_p$ for every prime $p$.

$\Leftarrow$ If $a_p\leq b_p$ for every prime $p$, then $b_p-a_p\geq 0$, and we can write $b=2^{b_2}3^{b_3}5^{b_5}\dots$ as \begin{align*}
    b &= 2^{a_2}3^{a_3}5^{a_5}\dots (2^{b_2-a_2}3^{b_3-a_3}5^{b_5-a_5}\dots) \\
      &= a(2^{b_2-a_2}3^{b_3-a_3}5^{b_5-a_5}\dots)
\end{align*}Since only finitely many $a_p, b_p$ are nonzero, it follows that only finitely many $b_p-a_p$ are nonzero, so $2^{b_2-a_2}3^{b_3-a_3}5^{b_5-a_5}\dots$ is a well defined integer. Thus, $a|b$. \qed

\textbf{Proposition (4.5):} Let $a=2^{a_2}3^{a_3}5^{a_5}\dots$ and $b=2^{b_2}3^{b_3}5^{b_5}\dots$ be positive integers. Then, let $d_p=\min(a_p,b_p)$ for every prime $p$. Then, $\gcd(a,b)=2^{d_2}3^{d_3}5^{d_5}\dots$

\underline{Example:} If $a=720=2^4\cdot 3^2 \cdot 5$ and $b=6400=2^2\cdot 5^2 \cdot 3^2\cdot 7=2^2\cdot 3^2\cdot 5^2\cdot 7$

Then, $d_2=2, d_3=2, d_5=1, d_7=0$, and all other $d_p=0$, so $\gcd(a,b)=2^2\cdot 3^2 \cdot 5 = 180$.

\textbf{Proof:} Let $d_p=\min(a_p,b_p)$ for every prime $p$, and let $d=2^{d_2}3^{d_3}5^{d_5}\dots$. We will show that $d=\gcd(a,b)$. By definition, $d_p\leq a_p$, and $d_p\leq b_p$ for every prime $p$, so by the previous proposition, $d|a$ and $d|b$. Thus, it is a common divisor. 

Let $e$ be an arbitrary common divisor. Then, $e=2^{e_2}3^{e_3}5^{e_5}\dots$ for some nonnegative integers $e_p$. Since $e|a$ and $e|b$, it follows by the previous proposition that $e_p\leq a_p$ and $e_p\leq b_p$ for every prime $p$. 

This implies that $e_p\leq \min(a_p,b_p)=d_p$ for every prime $p$, so by the previous proposition, $e|d$. Additionally, by def of divisibility $e\leq d$. Thus, $d$ is the greatest common divisor of $a$ and $b$. \qed

\newcommand*{\lcm}{\mathrm{lcm}}

\textbf{Definition (LCM):} The least common multiple of two positive integers $a,b$ is the smallest positive integer $m$ such that $a|m$ and $b|m$. We denote this as $\lcm(a,b)$.

\textbf{Proposition (4.5):} (continued) If $a=2^{a_2}3^{a_3}5^{a_5}\dots$ and $b=2^{b_2}3^{b_3}5^{b_5}\dots$ are positive integers, then let $m_p=\max(a_p,b_p)$ for every prime $p$. Then, $\lcm(a,b)=2^{m_2}3^{m_3}5^{m_5}\dots$

\textbf{Proof:} Let $m_p=\max(a_p,b_p)$ for every prime $p$, and let $m=2^{m_2}3^{m_3}5^{m_5}\dots$. We will show that $m=\lcm(a,b)$. By definition, $m_p\geq a_p$ and $m_p\geq b_p$ for every prime $p$, so by the previous proposition, $a|m$ and $b|m$. Thus, it is a common multiple.

\textbf{Corollary:} $\gcd(a,b)\cdot \lcm(a,b) = a\cdot b$ for all positive integers $a,b$.

\textbf{Proof:} Let $d_p=\min(a_p,b_p)$ and $m_p=\max(a_p,b_p)$ for every prime $p$. Then, \begin{align*}
    \gcd(a,b)\cdot \lcm(a,b) &= (2^{d_2}3^{d_3}5^{d_5}\dots)(2^{m_2}3^{m_3}5^{m_5}\dots) \\
                              &= 2^{d_2+m_2}3^{d_3+m_3}5^{d_5+m_5}\dots
\end{align*}Since $d_p=\min(a_p,b_p)$ and $m_p=\max(a_p,b_p)$, it follows that $d_p+m_p=a_p+b_p$ for every prime $p$. Thus, \begin{align*}
    \gcd(a,b)\cdot \lcm(a,b) &= 2^{a_2+b_2}3^{a_3+b_3}5^{a_5+b_5}\dots \\
                              &= (2^{a_2}3^{a_3}5^{a_5}\dots)(2^{b_2}3^{b_3}5^{b_5}\dots) \\
                              &= a\cdot b
\end{align*}\qed

\underline{Note:} This also implies that \begin{align*}
    \lcm(a,b) = \frac{a\cdot b}{\gcd(a,b)}
\end{align*}

\underline{Example:} (of Prop): If $a=720=2^4\cdot 3^2 \cdot 5$ and $b=6400=2^2\cdot 5^2 \cdot 3^2\cdot 7=2^2\cdot 3^2\cdot 5^2\cdot 7$, then $\lcm(a,b)=2^{4}3^{2}5^{2}\cdot 7 = 25200$.

\section{Lecture 13}

\subsection{We forgot a thing from Chapter 2:}

Apparently we forgot to prove a thing in chapter 2:

\textbf{Proposition (2.17):} Let $a,b,c$ be integers with $\gcd(a,b)=1$. If $a|c,b|c$, then $ab|c$.

\textbf{Proof:} Since $a|c, b|c$, we have that $c=ae, c=bf$ for integers $e,f$. Since $\gcd(a,b)=1$, there are integers $x,y$ such that \begin{align*}
    1=ax+by
\end{align*}

Multiplying by $c$ gives \begin{align*}
    c &= acx + bcy \\
      &= a(bf)x + b(ae)y \\
      &= ab(fx + ey)
\end{align*}
Thus, $ab|c$. \qed

\subsection{5.1 - A Puzzle}

\textbf{Proposition (5.1):} Let $k\geq 2$ be an integer, let $m$ be a positive integer. Then, $m$ is a $k$-th power of an integer iff all exponents in the prime factorization of $m$ are divisible by $k$. That is, if \begin{align*}
    m=2^{m_2}3^{m_3}5^{m_5}\dots
\end{align*}then $m$ is a $k$-th power of an integer iff $k|m_p$ for every prime $p$.

\textbf{Proof:} $\Rightarrow$ If $m$ is a $k$-th power of an integer, then there exists some integer $n$ such that $m=n^k$. Write $n=2^{n_2}3^{n_3}5^{n_5}\dots$. Then, \begin{align*}
    m &= n^k \\
      &= (2^{n_2}3^{n_3}5^{n_5}\dots)^k \\
      &= 2^{kn_2}3^{kn_3}5^{kn_5}\dots
\end{align*}By the fundamental theorem of arithmetic, the prime factorization of $m$ is unique, so $m_p=kn_p$ for every prime $p$. Since $n_p$ is an integer, it follows that $k|m_p$ for every prime $p$.

$\Leftarrow$ If $k|m_p$ for every prime $p$, then there exist integers $n_p$ such that $m_p=kn_p$ for every prime $p$. Then, \begin{align*}
    m &= 2^{m_2}3^{m_3}5^{m_5}\dots \\
      &= 2^{kn_2}3^{kn_3}5^{kn_5}\dots \\
      &= (2^{n_2}3^{n_3}5^{n_5}\dots)^k
\end{align*}Thus, $m$ is a $k$-th power of an integer. \qed

\underline{Puzzle:} We got a limerick puzzle:

\begin{center}
    There was a young lady named Chris,\\
    who, when asked her age, answered this\\
    Two-thirds of its square,\\
    Is a cube, I declare.\\
    Now what was the age of the miss?
\end{center}

Let $A$ be her age, eg \begin{align*}
    A=2^{a_1}3^{a_3}5^{a_5}\dots
\end{align*}

Two thirds of its square is \begin{align*}
    \frac{2}{3}A^2 &= \frac{2}{3}(2^{2a_2}3^{2a_3}5^{2a_5}\dots) \\
    &= 2^{2a_2+1}3^{2a_3-1}5^{2a_5}\dots
\end{align*}

Then, this must be a perfect cube, eg $2a_2+1$ must be divisible by 3, $2a_3-1$ must be divisible by 3, and $2a_i$ where $i\geq 5$ must be divisible by 3.

One possibility is $a_2=1, a_3=2, a_i=0$. This gives $A=2^1\cdot 3^2=18$. This is probably the most reasonable answer

Other possibilities? We have $3|2a_2+1, 3|2a_3-1, 3|2a_i$ for $i\geq 5$. Since $3|2a_i$, it follows that $3|a_i$ because the gcd of 3 and 2 is 1. 

Thus, $a_i=3k_i$ for some integer $k_i$.

Then, $3|(2a_2+1)$ implies that $2a_2+1=3k_2$ for some $k_2$. Note this implies $k_2$ must be odd! We can then write $k_2=2m_2+1$ for some integer $m_2$. Then, $a_2=3m_2+1$.

Similarly, $3|(2a_3-1)$ implies that $2a_3-1=3k_3$ for some integer $k_3$. Note this implies $k_3$ must be odd! We can then write $k_3=2m_3+1$ for some integer $m_3$. Then, $a_3=3m_3+2$.

Then, \begin{align*}
    A&=2^{3m_2+1}3^{3m_3+2}\prod_{i\geq 5} p_i^{3k_i}
    &=18(2^{m_2}3^{m_3}\prod_{i\geq 5} p_i^{k_i})^3
\end{align*} so the complete solution set is $18M^3$ where $M$ is an integer. The next smallest age, in fact, is 144. She could also be 1152 years old!

\subsection{5.2 - Irrationality Proofs}

\textbf{Definition (Rational):} $r\in \rr$ is rational if $\exists a,b\in \zz$ st $b\neq 0$ and $r=\frac{a}{b}$.

\textbf{Definition (Irrational):} $r\in \rr$ is irrational if it is not rational.

\newcommand*{\qq}{\mathbb{Q}}

\begin{aside}
The field of rational numbers is denoted $\qq$, and the field of irrational numbers is $\rr\setminus \qq$.

Also, because it is a field, we have that $\qq$ is closed under addition, subtraction, multiplication, and division (except by zero). 

Further, the field is dense, eg for any two rational numbers $r_1<r_2$, there exists a rational number $r$ such that $r_1<r<r_2$. In fact, there are infinitely many rational numbers between any two rational numbers.

\textbf{Proof:} Let $r_1=\frac{a}{b}, r_2=\frac{c}{d}$ where $a,b,c,d\in \zz$ and $b,d\neq 0$. Then, since $r_1<r_2$, it follows that $ad<bc$ (just cross multiply to see this).

Then, let $r=\frac{a+c}{b+d}$. We will show that $r_1<r<r_2$. First, \begin{align*}
    r-r_1 &= \frac{a+c}{b+d} - \frac{a}{b} \\
          &= \frac{(a+c)b - a(b+d)}{b(b+d)} \\
          &= \frac{cb - ad}{b(b+d)}
\end{align*}Since $ad<bc$, it follows that $cb-ad>0$, so $r-r_1>0$, so $r>r_1$. Next, \begin{align*}
    r_2-r &= \frac{c}{d} - \frac{a+c}{b+d} \\
          &= \frac{c(b+d) - (a+c)d}{d(b+d)} \\
          &= \frac{cb - ad}{d(b+d)}
\end{align*}Since $ad<bc$, it follows that $cb-ad>0$, so $r_2-r>0$, so $r<r_2$. Thus, there exists a rational number $r$ such that $r_1<r<r_2$. \qed

Any terminating decimal is rational.

\end{aside}

\section{Lecture 14}

\textbf{Definition (Rational):} If $r=\frac{a}{b}$, where $a,b$ are integers with $b\neq 0$, then $r$ is rational.

\underline{Notation:} \begin{itemize}
    \item $\zz$ is the set of all integers
    \item $\qq$ is the set of all rational numbers
    \item $\rr$ is the set of all real numbers
    \item $\rr\setminus \qq$ is the set of all irrational numbers
\end{itemize}

\textbf{Theorem:} If $r\in \qq$, we can always write \begin{align*}
    r=\frac{a}{b}
\end{align*}where $a,b$ are integers with $b>0$ and $\gcd(a,b)=1$.

\textbf{Proof:} Given any $r=\frac{a}{b}$ with $\gcd(a,b)=d$. We can write \begin{align*}
    r&=\frac{a}{b} \\
     &= \frac{a/d}{b/d}
\end{align*}and note that $a/d, b/d$ are integers with $b/d>0$ and $\gcd(a/d,b/d)=1$. \qed

\textbf{Theorem (5.2):} $\sqrt{2}$ is irrational.

\textbf{Proof:} On the contrary suppose it's rational. Then \begin{align*}
    \sqrt{2}&=\frac{a}{b}\\
    2&=\frac{a^2}{b^2} \\
    a^2&=2b^2
\end{align*}This shows that $2|a^2$. Since $2$ is prime, it follows that $2|a$. Thus, $a=2c$ for some $c\in \zz$. Then, \begin{align*}
    2b^2&=a^2 \\
         &= (2c)^2 \\
         &= 4c^2 \\
    b^2&=2c^2
\end{align*}which implies that $2|b^2$, so $2|b$. Thus, $a$ and $b$ are both even, which contradicts the fact that $\gcd(a,b)=1$. Thus, $\sqrt{2}$ is irrational. \qed

\textbf{Theorem (5.3):} Let $k\geq 2$ be an integer. Let $n$ be a positive integer that is not a $k$-th power of an integer. Then, $\sqrt[k]{n}$ is irrational.

\textbf{Proof:} We will show the contrapositive, 

\begin{aside}
The contrapositive states: If $\sqrt[k]{n}$ is rational, then $n$ is a $k$-th power of an integer.
\end{aside}

Let $\sqrt[k]{n}=\frac{a}{b}$ where $a,b$ are integers with $b>0$ and $\gcd(a,b)=1$. Then, \begin{align*}
    n&=\left(\frac{a}{b}\right)^k \\
     &= \frac{a^k}{b^k} \\
     &= 2^{k(a_2-b_2)}3^{k(a_3-b_3)}5^{k(a_5-b_5)}\dots
\end{align*}By the fundamental theorem of arithmetic, the prime factorization of $n$ is unique, so $n=2^{k(a_2-b_2)}3^{k(a_3-b_3)}5^{k(a_5-b_5)}\dots$. By the previous proposition, $n$ is a $k$-th power of an integer. \qed

\underline{Example:} $\sqrt[3]{9}, \sqrt[5]{8}$ are irrational. $2=\sqrt[4]{16}$ is rational. 

\subsection{5.3 - The Rational Root Theorem}

\textbf{Theorem (5.4 - Rational Root Theorem):} Let $P(x)=a_nx^n+a_{n-1}x^{n-1}+\cdots + a_1x + a_0$ be a polynomial with integer coefficients, where $a_n\neq 0, a_0\neq 0$. If $r=\frac{u}{v}$ is a reduced rational number such that $P(r)=0$, then $u|a_0$ and $v|a_n$.

\textbf{Proof:} Since $P(r)=0$, we have that \begin{align*}
    0&=a_n(\frac{u}{v})^n + a_{n-1}(\frac{u}{v})^{n-1} + \cdots + a_1(\frac{u}{v}) + a_0 \\
    \implies \\
    a_nu^n+a_{n-1}u^{n-1}v + \cdots + a_1uv^{n-1} + a_0v^n &= 0 \\
\end{align*}by multiplying both sides by $v^n$. Then, \begin{align*}
    a_0v^n&=-(a_nu^n+a_{n-1}u^{n-1}v + \cdots + a_1uv^{n-1}) \\
           &=-u(a_nu^{n-1}+a_{n-1}u^{n-2}v + \cdots + a_1v^{n-1})
\end{align*}Since $u$ divides the right hand side, it must divide the left hand side. Since $\gcd(u,v)=1$, it follows that $u|a_0$. Similarly, \begin{align*}
    a_nv^n&=-(a_{n-1}u^{n-1}v + \cdots + a_1uv^{n-1} + a_0v^n) \\
           &=-v(a_{n-1}u^{n-1} + \cdots + a_1u v^{n-2} + a_0v^{n-1})
\end{align*}Since $v$ divides the right hand side, it must divide the left hand side. Since $\gcd(u,v)=1$, it follows that $v|a_n$. \qed

\underline{Example:} Use the fact that \begin{align*}
    p(x)=2x^3+x^2-9
\end{align*}has a rational root to find all of its complex roots. If $r=\frac{u}{v}$ is a rational root, then $u|-9$ and $v|2$. Thus, $u=\pm 1,\pm 3, \pm 9$ and $v=\pm 1, \pm 2$. This implies the possible roots as \begin{align*}
    \pm 1,\pm 3, \pm 9, \pm \frac{1}{2}, \pm \frac{3}{2}, \pm \frac{9}{2}
\end{align*}Trying them all so we can see which one is a root, we see that $r=\frac{3}{2}$ is a root. Thus, $x-\frac{3}{2}$ is a factor, and \begin{align*}
    p(x) &= (x-\frac{3}{2})(2x^2+4x+6) \\
         &= 2(x-\frac{3}{2})(x^2+2x+3)
\end{align*}Then, we can use the quadratic formula to find the roots of the remaining quadratic. This is a way to factor a higher coefficient polynomial, basically!

\section{Lecture 15}

Recall last time we proved the rational root theorem, which states that if $P(x)=a_nx^n+a_{n-1}x^{n-1}+\cdots + a_1x + a_0$ is a polynomial with integer coefficients, where $a_n\neq 0, a_0\neq 0$, and $r=\frac{u}{v}$ is a reduced rational number such that $P(r)=0$, then $u|a_0$ and $v|a_n$.

Some applications

\underline{Example:} $\sqrt{2}$ is irrational. 

$\sqrt{2}$ is a root of the polynomial $p(x)=x^2-2$. If $\sqrt{2}$ were rational, then by the rational root theorem, the numerator of $\sqrt{2}$ in lowest terms would divide $-2$, and the denominator would divide $1$. Thus, the only possible rational roots are $\pm 1, \pm 2$. None of these are roots of $p(x)$, so $\sqrt{2}$ is irrational.

\underline{Example:} $\sqrt[k]{n}$ is irrational unless $n$ is a $k$-th power of an integer.

$\sqrt[k]{n}$ is a root of the polynomial $p(x)=x^k-n$. If $\sqrt[k]{n}$ were rational, then by the rational root theorem, the numerator of $\sqrt[k]{n}$ in lowest terms would divide $-n$, and the denominator would divide $1$. Thus, the only possible rational roots are $\pm 1, \pm n$. None of these are roots of $p(x)$ unless $n$ is a $k$-th power of an integer, so $\sqrt[k]{n}$ is irrational unless $n$ is a $k$-th power of an integer.

\underline{Example:} Take $p(x)=78x^3-574x^2+310x-45$. Say we know that the roots are approximately $0.26015\dots$, $0.3451\dots$, $6.40651\dots$. Are any of them rational?

Well, if any of them is rational, then the denominator is a divisor of 78. So if we multiply by 78, the root will become an integer. If we do that here, we get $20.2918, 27.0000, 499.7082$. Thus, $27$ is an integer, so $0.3451\dots$ is a rational root. We can then factor the polynomial with it. 

\subsection{5.4 - Pythagorean Triples}

\textbf{Definition (Pythagorean Triple):} A pythagorean triple is a triple of positive integers $(a,b,c)$ such that $a^2+b^2=c^2$, eg integer solutions to the pythagorean theorem. 

For instance, $(3,4,5)$ is a pythagorean triple since $3^2+4^2=5^2$. Also, $(5,12,13)$ is a pythagorean triple since $5^2+12^2=13^2$. Also, $(8,15,17)$ is a pythagorean triple since $8^2+15^2=17^2$.

Multiples too, eg $(6,8,10)$. 

\underline{Notes:} \begin{itemize}
    \item If $(a,b,c)$ is a pythagorean triple, then so is $(ka, kb, kc)$ for any positive integer $k$.
    \item If $(a,b,c)$ is a pythagorean triple, and $d=\gcd(a,b,c)$, then $\left(\frac{a}{d}, \frac{b}{d}, \frac{c}{d}\right)$ is a pythagorean triple.
\end{itemize}

\textbf{Definition (Primitive Pythagorean Triple):} A \underline{primitive} pythagorean triple is a pythagorean triple $(a,b,c)$ such that $\gcd(a,b,c)=1$.

\underline{Question:} Are there infinitely many primitive pythagorean triples? And can we generate them? Turns out the answer is yes for both

\textbf{Fibonacci's Method:} One way to do so: \begin{itemize}
    \item Start with any odd integer $a$. 
    \item Square it, and then write it as a sum of two consecutive integer. That is, write $a^2=k+(k+1)$ for some integer $k$.
    \item Then, we have that $a^2=k^2+(k+1)^2-(k^2+(k+1)^2-2k(k+1))=k^2+(k+1)^2-2k(k+1)=k^2+(k+1)^2-2k^2-2k=k^2+(k+1)^2-2k^2-2k=(k+1)^2-k^2$. Thus, $a^2=(k+1)^2-k^2$, so $(a, k, k+1)$ is a primitive pythagorean triple. (clear since the integers are consecutive so the gcd is 1)
\end{itemize}

\underline{Example:} \begin{itemize}
    \item $a=3$ yields $a^2=9=4+5$, so $k=4$, so we get the pythagorean triple $(3,4,5)$.
    \item $a=5$ yields $a^2=25=12+13$, so $k=12$, so we get the pythagorean triple $(5,12,13)$.
    \item $a=7$ yields $a^2=49=24+25$, so $k=24$, so we get the pythagorean triple $(7,24,25)$.
\end{itemize}

\textbf{Another Construction:} \begin{itemize}
    \item Start with a multiple of 4, eg $a=4k$
    \item Square it to $16k^2$ and then divide by 4 to get $4k^2$.
    \item Take the two odds on either side of $4k^2$, which are $4k^2-1$ and $4k^2+1$. Then, this is a primitive pythagorean triple since the three numbers are consecutive odd integers, so the gcd is 1.
\end{itemize}

\underline{Example:} \begin{itemize}
    \item $a=4$ yields $a^2=16$, dividing by 4 gives $4$, so we take the two odds on either side of $4$, which are $3$ and $5$, so we get the pythagorean triple $(4,3,5)$.
    \item $a=8$ yields $a^2=64$, dividing by 4 gives $16$, so we take the two odds on either side of $16$, which are $15$ and $17$, so we get the pythagorean triple $(8,15,17)$.
\end{itemize}

\textbf{Theorem (5.5):} Let $(a,b,c)$ be a primitive pythagorean triple. Then WLOG $c$ is odd and exactly one of $a,b$ is even. WLOG say $b$ is even. Then, there exists rprime integers $m<n$, one even, one odd, such that \begin{align*}
    a=n^2-m^2, b=2mn, c=n^2+m^2
\end{align*}Further any triple of this form is a primitive pythagorean triple.

\underline{Examples:} \begin{itemize}
    \item $n=3, m=2$ gives $a=9-4, b=2\cdot 3\cdot 2, c=9+4$, so we get the pythagorean triple $(5,12,13)$.
\end{itemize}

\section{Lecture 16}

We will prove 5.5. First, we will prove some related lemmas.

\textbf{Lemma (5.6):} Let $k\geq 2$ and let $a,b$ be rprime in $\zz$ such that $ab$ is a $k$-th power of an integer. Then, $a$ and $b$ are both $k$-th powers of integers.

\textbf{Proof:} Write the prime factorizations \begin{align*}
    a=p_1^{r_1}p_2^{r_2}\dots p_m^{r_m}, b=q_1^{s_1}q_2^{s_2}\dots q_n^{s_n}
\end{align*}Since $a,b$ are rprime, $p_i\neq q_j$ for all $i,j$. Then, \begin{align*}
    ab&=p_1^{r_1}p_2^{r_2}\dots p_m^{r_m}q_1^{s_1}q_2^{s_2}\dots q_n^{s_n}
\end{align*}is the prime factorization of ab into distinct primes. Since $ab$ is a $k$-th power of an integer, all exponents are multiples of $k$. Thus, $k|r_i$ for all $i$, and $k|s_j$ for all $j$. Thus, $a$ and $b$ are both $k$-th powers of integers. \qed

\textbf{Lemma (5.7):} \begin{enumerate}
    \item The square of an even number is divisible by 4
    \item The square of an odd number is one more than a multiple of 8
\end{enumerate}

\textbf{Proof:} \begin{enumerate}
    \item Let $a=2k$ for some integer $k$. Then, $a^2=4k^2$, which is divisible by 4.
    \item Let $a=2k+1$ for some integer $k$. Then, $a^2=(2k+1)^2=4k^2+4k+1=4k(k+1)+1$. Since one of $k,k+1$ is even, it follows that $4k(k+1)$ is divisible by 8. Thus, $a^2$ is one more than a multiple of 8.
\end{enumerate}\qed

\underline{Recall:}

\textbf{Theorem (5.5):} Let $(a,b,c)$ be a primitive pythagorean triple. Then WLOG $c$ is odd and exactly one of $a,b$ is even. WLOG say $b$ is even. Then, there exists rprime integers $m<n$, one even, one odd, such that \begin{align*}
    a=n^2-m^2, b=2mn, c=n^2+m^2
\end{align*}Further any triple of this form is a primitive pythagorean triple.

\textbf{Proof:} Let's start by proving the last assertion. That is, suppose $a^2=n^2-m^2, b=2mn, c=n^2+m^2$ for some rprime integers $m<n$, one even, one odd. Then, \begin{align*}
    a^2+b^2&=(n^2-m^2)^2+(2mn)^2 \\
        &=(n^4-2n^2m^2+m^4)+(4m^2n^2)\\
        &=n^4-2n^2m^2+m^4+4m^2n^2\\
        &=n^4+2n^2m^2+m^4 \\
        &=(n^2+m^2)^2 \\
\end{align*}

We will show that this is primitive. Suppose on the contrary that $p$ is a prime such that $p|a, p|b, p|c$. Then, note that $c=n^2+m^2$. Because one of $m,n$ is even, this is odd. Thus, $p\neq 2$. 

Then, $p$ divides $a+c=(n^2-m^2)+(n^2+m^2)=2n^2$ which implies $p|n^2\implies p|n$. 

Then, $p$ divides $c-a=(n^2+m^2)-(n^2-m^2)=2m^2$ which implies $p|m^2\implies p|m$.

However, this is a contradiction because $m,n$ are rprime. Thus, the triple is primitive.

Now, we will show the first assertion. Let $(a,b,c)$ be an arbitrary primitive pythagorean triple. We will show that WLOG $c$ is odd and exactly one of $a,b$ is even.

If $a,b$ are both even, then $c^2$ is divisible by 4, a contradiction. If $a,b$ are both odd, then $a^2\equiv 1\mod 8$ and $b^2\equiv 1\mod 8$, so $c^2=a^2+b^2\equiv 2\mod 8$, which is a contradiction. Thus, exactly one of $a,b$ is even. WLOG say $b$ is even. 

Then, $a^2+c^2=b^2$ is divisible by 4, so $a,c$ are both even or both odd. If $a,c$ are both even, then $b^2$ is divisible by 8, so $b$ is divisible by 4, so $a,b,c$ are all even, a contradiction. Thus, $a,c$ are both odd, so $c$ is odd. 

Write $b=2b_1$ for some integer $b_1$. Then, \begin{align*}
    4b_1^2=b^2=a^2-c^2=(c-a)(c+a)
\end{align*}which gives \begin{align*}
    b_1^2&=\frac{(c-a)(c+a)}{4} \\
         &=\left(\frac{c-a}{2}\right)\left(\frac{c+a}{2}\right)
\end{align*}These are both integers since $c$ is odd and $a$ is odd. 

We claim that $\gcd(\frac{c-a}{2}, \frac{c+a}{2})=1$. Suppose on the contrary that $p$ is a prime such that $p|\frac{c-a}{2}$ and $p|\frac{c+a}{2}$. Then, $p$ divides their sum, which is $c$, and their difference, which is $a$. Thus, $p|a, p|c$. Since $b^2=a^2+c^2$, it follows that $p|b^2$, so $p|b$. This contradicts the fact that $(a,b,c)$ is primitive. Thus, $\gcd(\frac{c-a}{2}, \frac{c+a}{2})=1$.

By the lemma 5.6, since $b_1^2=\left(\frac{c-a}{2}\right)\left(\frac{c+a}{2}\right)$ and $\gcd(\frac{c-a}{2}, \frac{c+a}{2})=1$, it follows that $\frac{c-a}{2}$ and $\frac{c+a}{2}$ are both perfect squares. Thus, there exist integers $m,n$ such that $\frac{c-a}{2}=m^2$ and $\frac{c+a}{2}=n^2$. 

Then, $c=n^2+m^2$ and $a=n^2-m^2$. Since $\gcd(\frac{c-a}{2}, \frac{c+a}{2})=1$, it follows that $\gcd(m^2,n^2)=1$, so $\gcd(m,n)=1$. 

Finally, since $b^2=a^2+c^2=(n^2-m^2)^2+(n^2+m^2)^2=4n^2m^2$, it follows that $b=2mn$. Additionally since $c$ is odd and $c=n^2+m^2$, it follows that one of $n,m$ is even and the other is odd. \qed

\section{Lecture 17}

\underline{Last Time:} We clasisfied all primitive solutions to $a^2+b^2=c^2$. That is, $a=n^2-m^2, b=2mn, c=n^2+m^2$ for $n\geq m$, $\gcd(m,n)=1$ for one even, one odd. 

What about solutions to \begin{align*}
    a^k+b^k=c^k
\end{align*}for $k>2$? Turns out, there are no solutions with positive integers $a,b,c$. This is \underline{Fermat's Last Theorem}, eventually proven by Andrew Wiles in 1995. 

\underline{Question:} Find all primitive Pythagorean Triples of the form $(21,b,c)$, if any exist at all. Then, if such a triple exists, it must satisfy that \begin{align*}
	21=n^2-m^2, b=2mn, c=n^2+m^2
\end{align*}

for some pair $(n,m)$ as in the Theorem. We need to write $21$ as a difference of two squares in order to achieve this result. This works with $n=11, m=10$, as $121-100=21$. Then, we have the triple $b=2(10)(11)$, and $c=11^2+10^2$, so $(21,220, 221)$ is a triple. 

Also, $n=5,m=2$ works, so we have $21, 20, 29$. 

\subsection{5.5 - Difference of Squares}

Given a positive integer $m$, how many ways (if any) can we write \begin{align*}
	m=x^2-y^2
\end{align*} as a difference of squares, where $x,y$ are integers?

\textbf{Theorem (5.8):} Let $m$ be a positive integer. Then, $m$ can be written as a difference of squares iff. $m$ is odd or $m$ is a multiple of $4$.

\textbf{Proof:} $\Leftarrow$ First note that \begin{align*}
	(n+1)^2-n^2=2n+1
\end{align*}So, any odd number can be written as a difference of squares. 

Next, note that \begin{align*}
	(n+1)^2-(n-1)^2=4n
\end{align*}This shows that any multiple of $4$ is a difference of two squares. 

$\Rightarrow$ Suppose $m=x^2-y^2$ is a difference of squares. Then, \begin{align*}
	m=(x-y)(x+y)
\end{align*}Note that the numbers $x-y,x+y$ are both either even or odd.  Thus, suppose $m=uv$ and $u,v$ have the same parity. Then, let $x=\frac{u+v}{2}$ and $y=\frac{u-v}{2}$. Then, $x^2-y^2=m$. Note that this happens only if $u,v$ have the same parity, so $m$ is either odd or a multiple of 4. \qed

\underline{Note:} So there is a 1-1 correspondence between factorizations $m=uv$, where $u,v$ have the same parity and $u\geq v$, and the ways to write $m=x^2-y^2$. 

\underline{Example:} Find all ways to write 21 as a difference of squares. Consider the factorizations of $21=uv$. \begin{itemize}
	\item $u=21,v=1$ gives $21=11^2-10^2$
	\item $u=7,v=3$ gives $21=5^2-2^2$
\end{itemize}

\underline{Example:} Find all ways to write 40 as a difference of squares. Consider the factorizations of $40=uv$. \begin{itemize}
	\item $u=40, v=1$ gives no valid solution because they are not the same parity 
	\item $u=20, v=2$ gives $40=11^2-9^2$
	\item $u=10, v=4$ gives $40=7^2-3^2$
	\item $u=8, v=5$ gives no valid solution because they are not the same parity. 
\end{itemize}

\underline{Example:} Find all ways to write 40 as a difference of squares. Consider the factorizations of $40=uv$. \begin{itemize}
	\item $u=40, v=1$ gives no valid solution because they are not the same parity 
	\item $u=20, v=2$ gives $40=11^2-9^2$
	\item $u=10, v=4$ gives $40=7^2-3^2$
	\item $u=8, v=5$ gives no valid solution because they are not the same parity. 
\end{itemize}

\underline{Example:} Find all primitive pythagorean triples with $a=45$. Note that we want to find $45=n^2-m^2$ as part of the pythagorean triple process. So write 45 as a difference of 2 squares! \begin{itemize}
	\item $u=45, v=1$ gives $n=23, m=22$. This gives us the valid pythagorean triple $(45, 1012, 1013)$
	\item $u=15, v=3$ gives $n=9, m=6$. However, because these 2 are not relatively prime, we cannot get a primitive pythagorean triple from this. But they do give a normal one. 
\end{itemize}

\section{Lecture 18}

Congruences

\subsection{6.1 - Definitions and Examples} 

Let $m$ be a positive integer. 

\textbf{Definition:} For $m$ a positive integer, two integers $a,b$ are congruent modulo $m$, denoted $a\equiv b\mod m$, if $m|a-b$.

\underline{Notation:} $a\equiv b\mod m$ means $a$ and $b$ are in the same residue class modulo $m$. THat is, \begin{align*}
    a\equiv b(\mod m)&\iff m|a-b \\
    &\iff a-b=km \text{ for some integer } k \\
    &\iff a=km+b \text{ for some integer } k
\end{align*}

\underline{Examples:} \begin{itemize}
    \item $8\equiv 0\mod 2$
    \item $13\equiv 1\mod 2$
    \item $13\equiv 1\mod 6$
    \item $5\equiv 17\mod 2$
\end{itemize}

\textbf{Proposition (6.2):} $a\equiv b(\mod m)\iff a=b+km$ for some integer $k$.

\textbf{Proof:} $\Rightarrow$ If $a\equiv b\mod m$, then $m|a-b$, so $a-b=km$ for some integer $k$, so $a=b+km$.

$\Leftarrow$ If $a=b+km$ for some integer $k$, then $a-b=km$, so $m|a-b$, so $a\equiv b\mod m$. \qed

\underline{Note:} Note that \begin{itemize}
    \item $a\equiv 0\mod 2\iff a$ is even
    \item $a\equiv 1\mod 2\iff a$ is odd
    \item \underline{and} every integer $a$ is congruent to either 0 or 1 mod 2. 
\end{itemize} Similarly $a\equiv 0\mod 3$ means $a$ is a multiple of 3, $a\equiv 1\mod 3$ means $a$ is one more than a multiple of 3, and $a\equiv 2\mod 3$ means $a$ is two more than a multiple of 3.

\textbf{Proposition (6.3):} If $a$ is an integer and $m$ is a positive integer then there is a unique integer $r$ with $0\leq r\leq m-1$ such that $a\equiv r\mod m$. Then, $r$ is called the \underline{least nonnegative residue of $a\mod m$}.

\textbf{Proof:} By the division algorithm, we can write $a=mq+r$ for some unique $q,r$ with $0\leq r\leq m-1$. Then, $a\equiv r\mod m$ by the previous proposition. Uniqueness follows from the fact that if $a\equiv r'\mod m$ for some $r'$ with $0\leq r'\leq m-1$, then $r\equiv r'\mod m$, so $m|r-r'$. Since $0\leq r,r'\leq m-1$, it follows that $r=r'$. \qed

\textbf{Proposition (6.4):} Let $a,b,c$ be integers and $M$ is a positive integer. Then \begin{enumerate}
    \item $a\equiv a\mod m$ (reflexivity)
    \item if $a\equiv b\mod m$, then $b\equiv a\mod m$ (symmetry)
    \item if $a\equiv b\mod m$ and $b\equiv c\mod m$, then $a\equiv c\mod m$ (transitivity)
\end{enumerate}

This proposition also says that congruence modulo $m$ is an equivalence relation on $\zz$. 

\textbf{Proof:}\begin{enumerate}
    \item $m$ divides $0=a-a$, so $a\equiv a\mod m$.
    \item Since $a\equiv b(\mod m)$, we know $m|a-b$, so $a-b=km$ for some integer $k$. Then, $b-a=-km$, so $m|b-a$, so $b\equiv a\mod m$.
    \item Since $a\equiv b\mod m$ and $b\equiv c\mod m$, we know $m|a-b$ and $m|b-c$, so $a-c=(a-b)+(b-c)$ is divisible by $m$, so $a\equiv c\mod m$.
\end{enumerate}\qed

\underline{Note:} It follows from everything we've done today that $a\equiv b\mod m\iff a,b$ have the same remainder when divided by $m$.

We'll do lots of arithmetic with congruences (\underline{modular arithmetic}), for instance $11=3\equiv 2\mod 12, 30+75\equiv 45\mod 60$, sometimes called \underline{clock arithmetic}

for the second one, observe that $30+75\equiv 30+15\mod 60$ since $75\equiv 15\mod 60$, and $30+15=45$.

\textbf{Proposition (6.5):} If $a\equiv b\mod m$ and $c\equiv d\mod m$, then \begin{enumerate}
    \item $a+c\equiv b+d\mod m$
    \item $a-c\equiv b-d\mod m$
    \item $ac\equiv bd\mod m$
\end{enumerate}

\textbf{Proof:} \begin{enumerate}
    \item Since $a\equiv b\mod m$ and $c\equiv d\mod m$, we know $m|a-b$ and $m|c-d$, so $a+c-(b+d)=(a-b)+(c-d)$ is divisible by $m$, so $a+c\equiv b+d\mod m$.
    \item Since $a\equiv b\mod m$ and $c\equiv d\mod m$, we know $m|a-b$ and $m|c-d$, so $a-c-(b-d)=(a-b)-(c-d)$ is divisible by $m$, so $a-c\equiv b-d\mod m$.
    \item Since $a\equiv b\mod m$ and $c\equiv d\mod m$, we know $m|a-b$ and $m|c-d$, so $ac-bd=a(c-d)+(a-b)d$ is divisible by $m$, so $ac\equiv bd\mod m$.
\end{enumerate}\qed

\section{Lecture 19}

\underline{Example:} "Simplify" $100+4\cdot 35\mod 33$. Then, \begin{align*}
    100+4\cdot35\equiv 1+4\cdot 2\mod 33
\end{align*}eg the least nonnegative residue of $100+4\cdot 35\mod 33$ is $9$.

\underline{Example:} $99^4\mod 101$ is $-2\mod 101$, so we have \begin{align*}
    99^4\equiv (-2)^4\mod 101
\end{align*}which is $16\mod 101$.

\textbf{Corollary (6.6):} If $a\equiv b\mod m$, then $a^n\equiv b^n\mod m$ for all positive integers $n$.

\textbf{Proof:} Use proposition 6.5's case 3 with induction. 

\underline{Warning:} If $a\equiv b\mod m$ and $n\equiv k\mod m$, then $a^n\equiv b^k\mod m$ is not necessarily true. For instance, $2\equiv 5\mod 3$ and $1\equiv 4\mod 3$, but $2^1=2\not\equiv 5^4=625\mod 3$.

We'll address exponents when we talk about Fermat's/Euler's theorem.

Another consequence of Prop 6.5 is that we can manipulate congruences by adding/subtracting/multiplying both sides by the same integer. That is, if we want to find $x$ such that \begin{align*}
    5x+1\equiv 8\mod 9
\end{align*}we can just solve mod 9 to get $x\equiv 5$. That is, \begin{align*}
    5x+1\equiv 8&\implies 5x\equiv 7\\
    &\implies 10x\equiv 14 \\
    &\implies x\equiv 5\mod 9
\end{align*}
\underline{Warning:} You cannot divide both sides of a congruence by the same integer in general. For instance, $2x\equiv 4\mod 6$ has the set $x\equiv 2\mod 6$ as ONE solution, but $x\equiv 5\mod 6$ is another solution that we got rid of by dividing by 2.

\textbf{Proposition (6.7):} If $ac\equiv bc\mod m$ and $\gcd(c,m)=1$, then $a\equiv b\mod m$. 

\textbf{Proof:} This follows from Proposition 6.8, which is

\textbf{Proposition (6.8):} If $ac\equiv bc(\mod m)$, and $gcd(c,m)=d$, then $a\equiv b\mod (m/d)$ AND that $a\equiv b+(m/d)k\mod m$ for some integer $k$ such that $0\leq k\leq d-1$. Essentially, it's saying that we can divide both sides by $c$ if we can also divide the modulus by the gcd of $c$ and $m$.

\underline{Example:} Suppose $10x\equiv 40\mod 55$. then, $\gcd(10,55)=5$, so we can divide both sides by 5 to get $2x\equiv 8\mod 11$. Since $\gcd(2,11)=1$, we can divide both sides by 2 to get $x\equiv 4\mod 11$.

\textbf{Proof:} Suppose $ac\equiv bc\mod m$ and $d=\gcd(c,m)$. Then, $ac=bc+km$ for some $k\in \zz$. Then, $ac-bc=km$, and $(a-b)c=km$. This implies $(a-b)\frac{c}{d}=k\frac{m}{d}$.

This shows that $m/d|\frac{c}{d}(a-b)$, and since $\gcd(\frac{c}{d}, \frac{m}{d})=1$, it follows that $m/d|a-b$, so $a\equiv b\mod (m/d)$.

So, $a=b+u(m/d)$ for some integer $u$. Write $u=dq+r$ with $0\leq r\leq d-1$ as in the division algorithm. Then, $a=b+(dq+r)(m/d)=b+rm+qdm\equiv b+rm\mod m$. Thus, $a\equiv b+rm\mod m$ for some integer $r$ such that $0\leq r\leq d-1$. \qed

\underline{Note:} Prime moduli are nice because they allow us to divide by any nonzero integer.

\textbf{Proposition (6.10):} If $p$ is a prime and $a,b$ are integers such that $ab\equiv 0\mod p$, then $a\equiv 0\mod p$ or $b\equiv 0\mod p$.

\textbf{Proof:} Suppose $ab\equiv 0\mod p$. If $a\equiv 0\mod p$, then we are done. If $a\not\equiv 0\mod p$, then $\gcd(a,p)=1$, so we can divide both sides of the congruence by $a$ to get $b\equiv 0\mod p$. \qed

\section{Lecture 20}

\textbf{Corollary (6.11):} Let $p$ be a prime. Then, the congruence $x^2\equiv 1\mod p$ has only the solutions $x\equiv 1\mod p$ and $x\equiv -1\mod p$.

\textbf{Proof:} Suppose $x^2\equiv 1\mod p$. Then, $x^2-1\equiv 0\mod p$, so $(x-1)(x+1)\equiv 0\mod p$. By the previous proposition, $x\equiv 1\mod p$ or $x\equiv -1\mod p$. \qed

\underline{Example:} Suppose $x^2\equiv 1\mod 8$. Then, we have the four solutions $x\equiv 1\mod 8$, $x\equiv 3\mod 8$, $x\equiv 5\mod 8$, and $x\equiv 7\mod 8$.

\underline{Example:} Suppose $x^2\equiv 1\mod 15$. Then, we have the four solutions $x\equiv 1\mod 15$, $x\equiv 4\mod 15$, $x\equiv 11\mod 15$, and $x\equiv 14\mod 15$.

This is because $x^2\equiv 1\mod 15$ is equivalent to $x^2\equiv 1\mod 3$ and $x^2\equiv 1\mod 5$. The first congruence has the solutions $x\equiv 1\mod 3$ and $x\equiv -1\mod 3$, and the second congruence has the solutions $x\equiv 1\mod 5$ and $x\equiv -1\mod 5$. By the Chinese Remainder Theorem, we can combine these to get the four solutions listed above.

\subsection{6.2 - Modular Exponentiation}

For a lot of cryptography, it's important to calculate $a^n\mod m$ for large $a,n,m$. 

\underline{Example:} Let's evaluate $5^{199}\mod 101$. We can keep squaring and reducing $\mod 101$, eg \begin{align*}
    5^2\equiv 25\mod 101\\
    5^4\equiv 25^2\equiv 625\equiv 19\mod 101\\
    \dots
    5^128\equiv 24\mod 101\\
\end{align*}Then, $199=128+64+4+2+1$ and so \begin{align*}
    5^199=5^{128}5^{64}5^4 5^2 5^1\equiv 24\cdot 78\cdot 19\cdot 25\cdot 5\equiv 81\mod 101
\end{align*}This is pretty efficient as it turns out! We just need to compute the mods in the powers of 2 and then break them down. 

\subsection{6.3 - Divisibility Tests}

Is $8394$ divisible by $2?$ is $9125$? It turns out that "$n$ is divisible by 2 iff its last digit is divisible by 2", as we all know. The decimal expansion of $8394$ is as follows: \begin{align*}
    8394&=8\cdot 10^3+3\cdot 10^2+9\cdot 10^1+4\cdot 10^0
    &\equiv 0 + 0 + 0 + 4\mod 2\\
    &\equiv 4\mod 2\\
    &\equiv 0\mod 2
\end{align*}

\textbf{Proposition (6.13):} $n$ is congruent to its last 2 digits mod 4, and is congruent to its last 3 digits mod 8.

\textbf{Proof:} This follows from the fact that $10^2\equiv 0\mod 4$ and $10^3\equiv 0\mod 8$.

\textbf{Proposition (6.14):} $n$ is congruent $\mod 3$ (or $\mod 9$) to the sum of its digits. Consequently, $n$ is divisible by 3 (or 9) iff the sum of its digits is divisible by 3 (or 9).

\textbf{Proof:} If $k\geq 1$, then $10^k\equiv 1\mod 3$ and $10^k\equiv 1\mod 9$. Thus, if $n=a_k10^k+a_{k-1}10^{k-1}+\dots + a_110+a_0$ is the decimal expansion of $n$, then \begin{align*}
    n&=a_k10^k+a_{k-1}10^{k-1}+\dots + a_110+a_0\\
    &\equiv a_k+a_{k-1}+\dots + a_1+a_0\mod 3
\end{align*}and \begin{align*}
    n&=a_k10^k+a_{k-1}10^{k-1}+\dots + a_110+a_0\\
    &\equiv a_k+a_{k-1}+\dots + a_1+a_0\mod 9
\end{align*}Thus, $n$ is divisible by 3 (or 9) iff the sum of its digits is divisible by 3 (or 9). \qed

\section{Lecture 21}

\underline{Last Time:} Divisibility tests: we did 3, 9, etc. 

Lastly, consider divisibility by 11. We know that $10\equiv -1\mod 11$, so $10^2\equiv 1\mod 11$, and generally \begin{align*}
    10^k\equiv (-1)^k\mod 11
\end{align*}

Suppose $n=a_k10^k+a_{k-1}10^{k-1}+\dots + a_110+a_0$ is the decimal expansion of $n$. Then, \begin{align*}
    n&=a_k10^k+a_{k-1}10^{k-1}+\dots + a_110+a_0\\
    &\equiv a_k(-1)^k+a_{k-1}(-1)^{k-1}+\dots + a_1(-1)+a_0\mod 11
\end{align*}Thus, $n$ is divisible by 11 iff the alternating sum of its digits is divisible by 11. For instance, $123456$ is divisible by 11 iff $1-2+3-4+5-6=-3$ is divisible by 11, which is not the case.

\textbf{Proposition (6.16):} $n$ is congruent mod 11 to the alternating sum of its digits. Consequently, $n$ is divisible by 11 iff the alternating sum of its digits is divisible by 11.

\textbf{Proof:} This follows from the fact that $10^k\equiv (-1)^k\mod 11$. \qed

\subsection{6.4 - Linear Convergence}

Given integers $a,b,m$ we want to solve the linear congruence \begin{align*}
    ax\equiv b\mod m
\end{align*} for $x$. We want $\pmod{m}$ solutions, so we usually give least nonnegative residues as solutions.

\underline{Example:} We could try all the possibilities, eg for \begin{align*}
    3x\equiv 2\pmod 5
\end{align*}

We can try $x=0,1,2,3,4$ to see if any of these work. We find that $x=4$ works, so the solution is $x\equiv 4\mod 5$.

\underline{Example:} We can try all the possibilities for \begin{align*}
    4x\equiv 2\pmod 10
\end{align*}has solutions $x=3,8,\dots$

\underline{Example:} We can try all the possibilities for \begin{align*}
    4x\equiv 3\pmod 10
\end{align*}has no solutions.

To study linear congruences, we will convert them to linear Diophantine equations. That is, we want to solve \begin{align*}
    ax\equiv b\mod m&\iff ax=b+km\text{ for some integer } k\\
    &\iff ax-km=b
\end{align*}for some integer $k$. This is a linear Diophantine equation in the variables $x$ and $k$. We know how to solve these from before, so we can solve linear congruences by converting them to linear Diophantine equations.

Recall that the diophantine equation \begin{align*}
    ax-my=b
\end{align*} has solutions iff $d=\gcd(a,m)|b$. If $d|b$, then there are infinitely many solutions, and the solutions are of the form \begin{align*}
    (x,y)=(x_0,y_0)+t\left(\frac{m}{d}, \frac{a}{d}\right)
\end{align*}for some integer $t$, where $(x_0,y_0)$ is a particular solution.

\section{Lecture 22}

\textbf{Theorem (6.18):} Let $m>0, a\neq 0\pmod m$. Then, the congruence $ax\equiv b\pmod m$ has a solution iff $d=\gcd(a,m)$ divides $b$. Additionally there are exactly $d$ distinct solutions up to $\pmod m$.

Moreover if $x_0$ is one solution, then the collection of all solutions $\pmod m$ are given by \begin{align*}
    x\equiv x_0+\frac{m}{d}\cdot k\pmod m, 0\leq k<d
\end{align*}

The solution $x_0$ can be taken to be any integer solution of \begin{align*}
    \frac{a}{d}x\equiv \frac{b}{d}\pmod{\frac{m}{d}}
\end{align*}

\textbf{Proof:} We know that \begin{align*}
    ax\equiv b\pmod m \iff \exists y ax-my=b
\end{align*}
Then, using Theorem 3.1, the diophantine equation has solutions iff $d|b$, and so the congruence has solutions iff that holds. 

If $d|b$, then all solutions of the Diophantine equation have the form \begin{align*}
    x=x_0-(\frac{-m}{d})k, y=y_0+\frac{a}{d}k\\
    x=x_0+\frac{m}{d}k, y=y_0+\frac{a}{d}k
\end{align*}

Here, $(x_0, y_0)$ is one solution of the Diophantine equation. Then, all solutions of $ax\equiv b\pmod m$ have the form \begin{align*}
    x\equiv x_0+\frac{m}{d}k\pmod m
\end{align*} for some integer $k$ (and $x_0$ is a solution of $ax\equiv b\pmod m$). Now, it only remains to show that $0\leq k<d$. Note that \begin{align*}
    x_0+\frac{m}{d}k\equiv x_0+\frac{m}{d}(k+d)\pmod m
\end{align*}so we can reduce $k$ mod $d$ to get $0\leq k<d$.

Lastly, if $x_0$ is any solution of $(\frac{a}{d}x_0)\equiv \frac{b}{d}\pmod{\frac{m}{d}}$, then \begin{align*}
    \frac{a}{d}x_0\equiv \frac{b}{d}\pmod{\frac{m}{d}}&\iff ax_0=b+mt\text{ for some integer } t\\
    &\iff ax_0\equiv b\pmod m
\end{align*}

\textbf{Corollary (6.19):} If $\gcd(a,m)=1$, then the congruence $ax\equiv b\pmod m$ has a unique solution $\pmod m$.

\underline{Example:} $4x\equiv 3\mod 10$ has no solutions since $\gcd(4,10)=2$ does not divide $3$.

\underline{Example:} $4x\equiv 2\mod 10$ has solutions since $\gcd(4,10)=2$ divides $2$. We can divide both sides by 2 to get $2x\equiv 1\mod 5$. 

$x_0=3$ solves this. All solutions must thus have the form \begin{align*}
    x\equiv 3+(10/2)k\mod 10\\
    x\equiv 3+5k\mod 10, 0\leq k<2
\end{align*}This gives us the two solutions $x\equiv 3\mod 10$ and $x\equiv 8\mod 10$.

\underline{Example:} $3x\equiv 12\pmod 18$ has solutions because $\gcd(3,18)=3$ divides $12$. 

Clearly $4$ is a solution, and all solutions must have the form \begin{align*}
    x\equiv 4+(18/3)k\mod 18\\
    x\equiv 4+6k\mod 18, 0\leq k<3
\end{align*}This gives us the three solutions $x\equiv 4\mod 18$, $x\equiv 10\mod 18$, and $x\equiv 16\mod 18$.

\underline{Example:} $100x\equiv 11\pmod{29}$. First, $100\equiv 13\pmod{29}$. Thus, we are solving $13x\equiv 11\pmod{29}$. 

Since $\gcd(13,29)=1$, there is a unique solution $\pmod{29}$. Then, we will try to solve \begin{align*}
    13x\equiv 11\pmod{29}&\iff 13x=11+29k\text{ for some integer } k\\
    &\iff 13x-29k=11
\end{align*}Use the extended Euclidean algorithm to solve this. We have \begin{align*}
    29&=13\cdot 2+3\\
    13&=3\cdot 4+1\\
    3&=1\cdot 3+0
\end{align*}Then, we can back substitute to get \begin{align*}
    1&=13-3\cdot 4\\
    &=13-4(29-13\cdot 2)\\
    &=9\cdot 13-4\cdot 29
\end{align*}Thus, $11=11\cdot 1=11(9\cdot 13-4\cdot 29)=99\cdot 13-44\cdot 29$. This gives us the solution $x=99\equiv 12\mod 29$. That is, $100*12\equiv 11\mod 29$.

\subsection{Multiplicative Inverses}

\textbf{Definition (Multiplicative Inverse):} We say that integers $a,b$ are multiplicative inverses $\mod m$ if $ab\equiv 1\mod m$.

Then we will write that \begin{align*}
    a^{-1}\equiv b\mod m, \text{ or } b^{-1}\equiv a\mod m
\end{align*}

\underline{Example:} $(13)(9)\equiv 1\pmod{29}$ so $13^{-1}\equiv 9\mod 29$ and $9^{-1}\equiv 13\mod 29$.

\underline{Example:} In $\pmod 7$, $2\cdot 4\equiv 1\pmod 7$ so $2^{-1}\equiv 4\mod 7$ and $4^{-1}\equiv 2\mod 7$.

Note that the inverse of $a\pmod m$ is a solution to \begin{align*}
    ax\equiv 1\mod m
\end{align*}

\section{Lecture 23}

\subsection{Multiplicative Inverses $\pmod m$}

Recall that for integers $a,b$ and a positive integer $m$, we say that $a$ and $b$ are multiplicative inverses $\mod m$ if $ab\equiv 1\mod m$.

\underline{Example:} $\pmod 10$ \begin{align*}
    1\cdot 1\equiv 1\mod 10\implies 1^{-1}\equiv 1\mod 10\\
    3\cdot 7\equiv 1\mod 10\implies 3^{-1}\equiv 7\mod 10\\
    9\cdot 9\equiv 1\mod 10\implies 9^{-1}\equiv 9\mod 10
\end{align*}
here, 0,2,4,5,6,8 do not have inverses $\mod 10$. This is because $\gcd(0,10)=10$, $\gcd(2,10)=2$, $\gcd(4,10)=2$, $\gcd(5,10)=5$, $\gcd(6,10)=2$, and $\gcd(8,10)=2$ are not equal to 1.

\underline{Note:} The integers that have multiplicative inverses $\mod m$ are precisely those that are rprime to $m$.

\textbf{Corollary (6.21):} The integer $a$ has an inverse $\pmod m$ iff $\gcd(a,m)=1$.

Furthermore, in this case, the inverse is unique $\pmod m$.

\textbf{Proof:} In order to find the inverse of $a\pmod m$, we need to solve \begin{align*}
    ax\equiv 1\mod m
\end{align*}By Theorem 6.18, this has a solution iff $\gcd(a,m)=1$. 

Then, suppose $x_0$ is a solution. Then, all solutions have the form \begin{align*}
    x\equiv x_0+(m/1)k\mod m\\
    x\equiv x_0\mod m
\end{align*}so the solution is unique $\mod m$. \qed

\underline{Example:} In a previous example we found that $13(9)+29(-4)=1$, so $13(9)\equiv 1\mod 29$, so $13^{-1}\equiv 9\mod 29$.

\underline{Example:} Find the inverse of $49\pmod{100}$. We can use the extended Euclidean algorithm to find integers $x,y$ such that $49x-100y=1$. We have \begin{align*}
    100&=49\cdot 2+2\\
    49&=2\cdot 24+1\\
    2&=1\cdot 2+0
\end{align*}Then, we can back substitute to get \begin{align*}
    1&=49-2\cdot 24\\
    &=49-24(100-49\cdot 2)\\
    &=49\cdot 49-24\cdot 100
\end{align*}Thus, $49(49)\equiv 1\mod 100$, so $49^{-1}\equiv 49\mod 100$.

Additionally, this inverse is unique $\mod 100$, so the only solution to $49x\equiv 1\mod 100$ is $x\equiv 49\mod 100$.

\begin{aside}
Suppose $x+1|m$. Then, is the inverse of $x\mod m$ always equal to itself if $\gcd(x,m)=1$? For instance, $49+1=50|100$, and $49^{-1}\equiv 49\mod 100$.

No, because $x=2, m=21$ is a counterexample. But what if $\frac{m}{x+1}|x-1$? That is, if $x+1$ divides $m$, and the ratio $\frac{m}{x+1}$ divides $x-1$. 

\textbf{Proof:} Suppose $x+1|m$, and $\gcd(x,m)=1$. (That is, $x$ has an inverse $\mod m$). Additionally, let $\frac{m}{x+1}|x-1$. That is, $x-1=k\cdot \frac{m}{x+1}$ for some integer $k$.

Then, $m=k(x+1)$ for some integer $k$ by assumption. Then, we aim to show $x^2\equiv 1\pmod m$.

Note that $x^2-1=(x-1)(x+1)$. Note that $x+1|(x-1)(x+1)$, so $k(x+1)|k(x-1)(x+1)$, so $m|k(x-1)(x+1)$, so $m|k(x^2-1)$. We aim to show that $k|(x-1)$. 

However, this is immediate since $x-1=k\cdot \frac{m}{x+1}$. Thus, $k|(x-1)$, so $m|k(x^2-1)$ implies $m|(x^2-1)$, so $x^2\equiv 1\mod m$. Thus, $x^{-1}\equiv x\mod m$. \qed

Furthermore, all families $x=kt+1, m=k(kt+2)$ for some $k>0, t$ satisfies this. For example, if $k=2, t=3$, then $x=2*3+1=7$ and $m=2(6+2)=16$. Then, $7^{-1}\equiv 7\mod 16$.
\end{aside}

\underline{Note:} Using the inverse to solve $ax\equiv b\pmod m$: If $a$ has an inverse $\pmod m$, then \begin{align*}
    ax\equiv b\pmod m\\
    a^{-1}ax\equiv a^{-1}b\mod m\\
    x\equiv a^{-1}b\mod m
\end{align*}is the unique solution $\mod m$.

\subsection{6.5 - Chinese Remainder Theorem}

1700 years ago, Sun Tzu asks for a number $x$ that leaves a remainder of 2 when divided by 3, a remainder of 3 when divided by 5, and a remainder of 2 when divided by 7. That is, we want to solve the system of congruences \begin{align*}
    x&\equiv 2\mod 3\\
    x&\equiv 3\mod 5\\
    x&\equiv 2\mod 7
\end{align*}

\section{Lecture 24}

\underline{Goal:} We aim to find all $c$ such that $x\equiv c\pmod {mn}$, where \begin{align*}
    x&\equiv a\mod m\\
    x&\equiv b\mod n
\end{align*}

Here's an approach that generalizes well:

\underline{Example:} Take \begin{align*}
    \begin{cases}
        x&\equiv 1\pmod 3\\
        x&\equiv 5\pmod 7
    \end{cases}
\end{align*}

The first congruence says that $x=1+3t$ for some $t$. Let's insert this into the second congruence, eg \begin{align*}
    x&\equiv 5\pmod 7\\
    1+3t&\equiv 5\pmod 7\\
    3t&\equiv 4\pmod 7\\
    t&\equiv 6\pmod 7
\end{align*}so $t=6+7u$ for some $u$. Then, $x=1+3t=1+3(6+7u)=19+21u$. Thus, the solutions to the system of congruences are $x\equiv 19\mod 21$.

\underline{Example:} Solve \begin{align*}
    \begin{cases}
        x\equiv 1\pmod 2\\
        2x\equiv 1\pmod 5\\
        x\equiv 7\pmod 9
    \end{cases}
\end{align*}

Note that this system yields \begin{align*}
    \begin{cases}
        x&\equiv 1\pmod 2\\
        x&\equiv 3\pmod 5\\
        x&\equiv 7\pmod 9
    \end{cases}
\end{align*}Then, $x\equiv 1\pmod 2\implies x=1+2t$ for some $t$. Inserting this into the second congruence gives us \begin{align*}
    1+2t&\equiv 3\pmod 5\\
    2t&\equiv 2\pmod 5\\
    t&\equiv 1\pmod 5
\end{align*}so $t=1+5u$ for some $u$. Then, $x=1+2t=1+2(1+5u)=3+10u$. Inserting this into the third congruence gives us \begin{align*}
    3+10u&\equiv 7\pmod 9\\
    10u&\equiv 4\pmod 9\\
    u&\equiv 4\pmod 9
\end{align*}so $u=4+9v$ for some $v$. Then, $x=3+10u=3+10(4+9v)=43+90v$. Thus, the solutions to the system of congruences are $x\equiv 43\mod 90$.

\textbf{Theorem (6.22):} (Chinese Remainder Theorem, 1st version): If $m$ and $n$ are relatively prime, then the system of congruences \begin{align*}
    x&\equiv a\mod m\\
    x&\equiv b\mod n
\end{align*}has a unique solution $\mod mn$.

\underline{Note:} We know that \begin{align*}
    x=a+mt \text{ for some integer } t\\
    \implies a+mt\equiv b\mod n\\
    \implies mt\equiv b-a\mod n\\
    \implies t\equiv m^{-1}(b-a)\mod n
\end{align*}Thus, $t$ is uniquely determined $\mod n$, so $x=a+mt$ is uniquely determined $\mod mn$.

\textbf{Proof:} (Existence): Let $t$ be an integer with the property that $t$ satisfies $mt\equiv b-a\mod n$. Since $\gcd(m,n)=1$, $m$ has an inverse $\mod n$, so such a $t$ exists. 

Let $x=a+mt$. Then, we claim $x$ is a solution of the system. Note that $\pmod m$, $x\equiv a+mt\equiv a\pmod m$. Also, $\pmod n$, $x\equiv a+mt\equiv a+(b-a)\equiv b\pmod n$. Thus, $x$ is a solution of the system.

(Uniqueness): Suppose $x_1,x_2$ are two solutions. Then, \begin{align*}
    x_1&\equiv a\mod m\\
    x_2&\equiv a\mod m\\
    x_1&\equiv b\mod n\\
    x_2&\equiv b\mod n
\end{align*}But this tells us that \begin{align*}
    x_1\equiv x_2\mod m\\
    \implies m|x_1-x_2\\
\end{align*}and similarly $n|x_1-x_2$. Since $\gcd(m,n)=1$, it follows that $mn|x_1-x_2$, so $x_1\equiv x_2\mod mn$ by Prop 2.17. \qed

\textbf{Theorem:} (Chinese Remainder Theorem, 2nd version): Let $m_1,m_2,\dots, m_k$ be pairwise relatively prime positive integers. Let $M=m_1m_2\cdots m_k$. Then, the system of congruences \begin{align*}
    x&\equiv a_1\mod m_1\\
    x&\equiv a_2\mod m_2\\
    &\vdots\\
    x&\equiv a_k\mod m_k
\end{align*}has a unique solution $\mod M$.

\textbf{Proof:} We will prove this by induction on $k$, the number of congruences. 

\underline{Note:} Also, the CRT says that a congruence with a composite modulus is equivalent to a \underline{system} of congruences with rprime moduli, which are usually the factors of the composite moduli. 

\underline{Example:} Suppose we have the problem $x\equiv 33\pmod{90}$. Then, we can decompose this to \begin{align*}
    x&\equiv 33\mod 2\\
    x&\equiv 33\mod 5\\
    x&\equiv 33\mod 9
\end{align*}, which is in turn equivalent to \begin{align*}
    x&\equiv 1\mod 2\\
    x&\equiv 3\mod 5\\
    x&\equiv 6\mod 9
\end{align*}

\section{Lecture 25}

Exam on Monday, up to and including 6.5.

\underline{Example:} Find all $\pmod{55}$ solutions to $x^2\equiv 1\pmod{55}$. 

\begin{align*}
    x^2\equiv 1\mod 55&\iff \begin{cases}
    x^2\equiv 1\mod 5\\
    x^2\equiv 1\mod 11
    \end{cases}\\
    &\iff \begin{cases}
        x\equiv \pm 1\mod 5\\
        x\equiv \pm 1\mod 11
    \end{cases}
\end{align*}This gives us 4 cases. Working through them, \begin{enumerate}
    \item $\begin{cases}
        x\equiv 1\mod 5\\
        x\equiv 1\mod 11
    \end{cases}\implies x\equiv 1\mod 55$
    \item $\begin{cases}
        x\equiv 1\mod 5\\
        x\equiv -1\mod 11
    \end{cases}\implies x\equiv 21\mod 55$
    \item $\begin{cases}
        x\equiv -1\mod 5\\
        x\equiv 1\mod 11
    \end{cases}\implies x\equiv 34\mod 55$
    \item $\begin{cases}
        x\equiv -1\mod 5\\
        x\equiv -1\mod 11
    \end{cases}\implies x\equiv 54\mod 55$
\end{enumerate}
Thus we have the four solutions $x\equiv 1\mod 55$, $x\equiv 21\mod 55$, $x\equiv 34\mod 55$, and $x\equiv 54\mod 55$.

\underline{Note:} Generally, if $m=p_1^{r_1}p_2^{r_2}\cdots p_k^{r_k}$, then $x^2\equiv 1\mod m$ is equivalent to the system of congruences \begin{align*}
    x^2&\equiv 1\mod p_1^{r_1}\\
    x^2&\equiv 1\mod p_2^{r_2}\\
    &\vdots\\
    x^2&\equiv 1\mod p_k^{r_k}
\end{align*}Also, genearlly, $a\equiv b\pmod m$ is equivalent to the system of congruences \begin{align*}
    a&\equiv b\mod p_1^{r_1}\\
    a&\equiv b\mod p_2^{r_2}\\
    &\vdots\\
    a&\equiv b\mod p_k^{r_k}
\end{align*}

\subsection{8.1 - Fermat's Theorem}

\textbf{Theorem (8.1):} (Fermat's Theorem): Let $p$ be a prime. Then for every integer $b$, \begin{align*}
    b^p\equiv b\pmod p
\end{align*}

Additionally if $b\not\equiv 0\pmod p$, then \begin{align*}
    b^{p-1}\equiv 1\pmod p
\end{align*}

\underline{Note:} the parts of Fermat's theorem imply each other, since if $b\not\equiv 0\pmod p$, then $b^p\equiv b\mod p$ implies $b^{p-1}\equiv 1\mod p$ by multiplying by the inverse of $b\mod p$. Additionally it goes the other way around because if $b\not\equiv 0\pmod p$, then $b^{p-1}\equiv 1\mod p$ implies $b^p\equiv b\mod p$ by multiplying by $b$.

\underline{Note:} Consider $b=2$. Then, \begin{align*}
    2^2\equiv 4\equiv 2\pmod 2\\
    2^3\equiv 8\equiv 2\pmod 3\\
    2^4\equiv 16\equiv 0\not\equiv 2\pmod 4\\
    2^5\equiv 32\equiv 2\pmod 5\\
\end{align*}

\underline{Example:} Simplify $3^{24}\pmod 7$. Since $3^6\equiv 1\mod 7$ by Fermat's theorem, we have $3^{24}=(3^6)^4\equiv 1^4\equiv 1\mod 7$.

\underline{Example:} Simplify $7^{182}\pmod{19}$. Fermat says that $7^{18}\equiv 1\mod 19$, so $7^{182}=7^{18\cdot 10+2}=(7^{18})^{10}\cdot 7^2\equiv 1^{10}\cdot 49\equiv 11\mod 19$.

\textbf{Corollary (8.2):} Let $p$ be prime and $b\not\equiv 0\pmod p$. If $x\equiv y\pmod{p-1}$ then $b^x\equiv b^y\pmod p$.

\textbf{Proof:} Suppose $x\equiv y\pmod{p-1}$, so that \begin{align*}
    x=y+(p-1)k\text{ for some integer } k
\end{align*}

Then $b^x=b^{y+(p-1)k}=(b^{p-1})^kb^y\equiv 1^kb^y\equiv b^y\mod p$. \qed

So when working $\mod p$, we can reduce exponents $\mod p-1$.

\underline{Example:} Simplify $\pmod 5$. \begin{align*}
    127^{53}+58^{70}&\equiv 2^{53}+3^{70}\mod 5\\
    &\equiv 2^1+3^2\mod 5\\
    &\equiv 2+9\equiv 1\mod 5
\end{align*}

\underline{Note:} Fermat's Theorem says that if $p$ is prime and $b\not\equiv 0\pmod p$, then \begin{align*}
    b^{p-1}\equiv 1\mod p
\end{align*} So if $n$ is a positive integer and we find a $b\not\equiv 0\mod p$ such that $b^{n-1}\not\equiv 1\mod p$, then $n$ is not prime. For instance, $2^4=16\equiv 1\mod 5$, so this does not show that 5 is not prime. However, $2^4=16\equiv 0\not\equiv 1\mod 4$, so this shows that 4 is not prime.

Also, if $2^{n-1}\not\equiv 1\pmod n$, then $n$ is composite. 

\underline{Example:} Let's show $n=133$ is composite by evaluating $2^{132}\mod 133$. We have that \begin{align*}
    2^2\equiv 4\mod 133\\
    2^4\equiv 16\mod 133\\
    2^8\equiv 256\equiv 123\mod 133\\
    2^{16}\equiv 123^2\equiv 100\mod 133\\
    \vdots \\
    2^{128}\equiv 4\mod 133
\end{align*}

Thus, $2^{132}=2^{128}\cdot 2^4\equiv 4\cdot 16\equiv 64\not\equiv 1\mod 133$, so $133$ is composite.

\section{Lecture 26}

Exam

\section{Lecture 27}

\textbf{Lemma (8.4):} Let $p$ be a prime and $b\not \equiv 0\pmod p$. Then the set of numbers \begin{align*}
    \{b, 2b, 3b, \dots, (p-1)b\}\pmod p =\{1, 2, 3, \dots, p-1\}\pmod p
\end{align*}

\textbf{Example:} If we have $p=11, b=3$, then \begin{align*}
    \{b, 2b, 3b, \dots, 10b\} &=\{3, 6, 9, 1, 4, 7, 10, 2, 5, 8\}\\
    &=\{1, 2, 3, 4, 5, 6, 7, 8, 9, 10\}
\end{align*}

\textbf{Proof:} Note that if $1\leq k\leq p-1$ then $kb\not\equiv 0\pmod p$.

To see the elements of $\{b, 2b, \dots, (p-1)b\}$ are different, suppose $ib\equiv jb\pmod p$ for some $1\leq i,j\leq p-1$. Then, multiply by $b^{-1}$ to get $i\equiv j\pmod p$. Since $1\leq i,j\leq p-1$, this implies $i=j$. Thus, the elements of $\{b, 2b, \dots, (p-1)b\}$ are different. Since there are $p-1$ elements in $\{b, 2b, \dots, (p-1)b\}$ and $p-1$ elements in $\{1, 2, \dots, p-1\}$, the two sets must be equal. \qed

\textbf{Theorem (Fermat's (Little) Theorem):} Let $p$ be a prime. Then, \begin{enumerate}
    \item For any integer $b$ \begin{align*}
        b^p\equiv b\pmod p
    \end{align*}
    \item If $b\not\equiv 0\pmod p$ then \begin{align*}
        b^{p-1}\equiv 1\pmod p
    \end{align*}
\end{enumerate} 

\textbf{Proof:} We know that these are equivalent to each other. Then, we will prove b. 

Let $b\not\equiv 0\pmod p$. Then by lemma \begin{align*}
    \{b, 2b, 3b, \dots, (p-1)b\} &=\{1, 2, 3, \dots, p-1\}\\
    \implies b\cdot 2b\cdot 3b\cdots (p-1)b&=1\cdot 2\cdot 3\cdots (p-1)\\
    b^{p-1}\cdot (p-1)!&=(p-1)!
\end{align*}Since $\gcd((p-1)!, p)=1$, we can divide both sides by $(p-1)!$ to get $b^{p-1}\equiv 1\mod p$. \qed

Let $\phi(n)$ denote the number of elements in the set $T=\{1\leq k\leq n: \gcd(k,n)=1\}$. For instance, $\phi(10)=4$ since the elements of $T$ are 1,3,7,9.

Then \begin{align*}
    \phi(p)=p-1\text{ for prime } p
    \phi(2)=1\\
    \phi(3)=2\\
    \phi(5)=4\\
    \phi(7)=6\\
\end{align*}for primes. But $\phi(4)=2$

\subsection{8.9 - Euler's Theorem}

\underline{Fact:} The product of two or more invertibles $\pmod n$ is also invertible. Additionally, the inverse of $ab$ is $(ab)^{-1}=b^{-1}a^{-1}$. 

Suppose $\gcd(b,n)=1$ so that $b$ is invertible. Let $bT=\{b a\pmod n: a\in T\}$. Then, $bT$ is the set of invertibles $\pmod n$. Thus, $|bT|=|T|=\phi(n)$.

\textbf{Theorem:} $bT=T$. 

\textbf{Proof:} Everything in $bT$ is a product of invertibles, and thus is invertible, and $bT\subseteq T$. 

Then, given any $a\in T$, we can write \begin{align*}
    a\equiv b(b^{-1}a)\mod n
\end{align*}Since $b^{-1}a\in T$, we have $b^{-1}a\equiv bT$ so $a\equiv b(b^{-1}a)\equiv bT$. Thus, $T\subseteq bT$. \qed

\textbf{Theorem (8.9):} (Euler's Theorem): Suppose $n$ positive and $\gcd(b,n)=1$, then \begin{align*}
    b^{\phi(n)}\equiv 1\pmod n
\end{align*}

\textbf{Proof:} By the lemma, $T=bT$. Thus, \begin{align*}
    \prod_{a\in T}a&\equiv\prod_{x\in bT}x\\
    &\equiv \prod_{a\in T}ba\pmod n\\
    (\prod_{a\in T}a)&\equiv b^{\phi(n)}(\prod_{a\in T}a)\mod n
\end{align*}Since $\prod_{a\in T}a$ is a product of invertibles, it is invertible mod $n$. Multiply by $(\prod_{a\in T}a)^{-1}$ to get $1\equiv b^{\phi(n)}\mod n$. \qed

\underline{Remark:} If $n=p$ a prime, then $\phi(p)=p-1$ and this is Fermat's

\underline{Example:} $n=14$ has $\phi(14)=6$ since the elements of $T$ are 1,3,5,9,11,13. Then, if $\gcd(b,14)=1$, then $b^6\equiv 1\mod 14$.

\underline{Example:} Reduce $5^{362}\pmod{14}$. Then, \begin{align*}
    5^{362}&\equiv 5^{6\cdot 60+2}\mod 14\\
    &\equiv (5^6)^{60}\cdot 5^2\mod 14\\
    &\equiv 1^{60}\cdot 25\mod 14\\
    &\equiv 11\mod 14
\end{align*}

\section{Lecture 28}

\underline{Recall:} Euler's theorem says that if $n$ is a positive integer and $\gcd(b,n)=1$, then $b^{\phi(n)}\equiv 1\mod n$, where $\phi(n)=|\{1\leq k\leq n: \gcd(k,n)=1\}|$.

\textbf{Corollary (8.1):} Let $n$ be a positive integer and suppose $\gcd(b,n)=1$. Then, if $x\equiv y\pmod{\phi(n)}$ then \begin{align*}
    b^x\equiv b^y\pmod n
\end{align*}

\textbf{Proof:} $x\equiv y\pmod{\phi(n)}$ implies $x=y+\phi(n)k$ for some integer $k$. Then, \begin{align*}
    b^x&=b^{y+\phi(n)k}\\
    &=(b^{\phi(n)})^kb^y\\
    &\equiv 1^kb^y\mod n\\
    &\equiv b^y\mod n
\end{align*} \qed

\underline{Example:} Reduce $47^{50}\pmod{15}$. This is \begin{align*}
    47^{50}&\equiv 2^{50}\mod 15\text{ since } 47\equiv 2\mod 15\\
    &\equiv 2^{2}\mod 15\text{ since } \phi(15)=8, 50\pmod 8=2\\
    &\equiv 4\mod 15
\end{align*}

To effectively use Euler's theorem, we need to be able to compute $\phi(n)$ efficiently.

\textbf{Proposition (8.7):} If $p$ is a prime and $k\geq 1$, then \begin{align*}
    \phi(p^k)=p^k-p^{k-1}=p^{k-1}(p-1)
\end{align*}

\textbf{Proof:} There are $p^k$ integers $a$ such that $1\leq a\leq p^k$. Then, note that $\gcd(a,p^k)\neq 1\iff p|a\iff a=pt$ for some integer $t$. There are $p^{k-1}$ integers $a$ such that $1\leq a\leq p^k$ and $p|a$, since there are $p^{k-1}$ integers $t$ such that $1\leq t\leq p^{k-1}$. Thus, there are $p^k-p^{k-1}$ integers $a$ such that $1\leq a\leq p^k$ and $\gcd(a,p^k)=1$. \qed

\textbf{Proposition (8.6):} If $\gcd(m,n)=1$, then $\phi(mn)=\phi(m)\phi(n)$.

\textbf{Proof:} Use the 2 previous propositions.

\underline{Example:} $\phi(100)=\phi(2^2\cdot 5^2)=\phi(2^2)\cdot \phi(5^2)=(2^2-2^1)(5^2-5^1)=40$.

\underline{Example:} $\phi(120)=\phi(2^3\cdot 3 \cdot 5)=\phi(2^3)\cdot \phi(3)\cdot \phi(5)=(2^3-2^2)(3-1)(5-1)=32$.

\textbf{Theorem (8.8):} Let $n=p_1^{a_1}p_2^{a_2}p_5^{a_5}\dots p_r^{a_r}$ be the prime factorization of $n$. Then, \begin{enumerate}
    \item $\phi(n)=\prod_{i=1}^r p_i^{a_i}-p_i^{a_i-1}$
    \item $\phi(n)=n\prod_{i=1}^r (1-\frac{1}{p_i})=n\prod_{p|n, p\text{ prime}}(1-\frac{1}{p})$
\end{enumerate}

\textbf{Proof:} 1. \begin{align*}
    \phi(n)&=\phi(p_1^{a_1}p_2^{a_2}p_5^{a_5}\dots p_r^{a_r})\\
    &=\phi(p_1^{a_1})\cdot \phi(p_2^{a_2})\cdot \phi(p_5^{a_5})\cdots \phi(p_r^{a_r})\\
    &=(p_1^{a_1}-p_1^{a_1-1})(p_2^{a_2}-p_2^{a_2-1})(p_5^{a_5}-p_5^{a_5-1})\cdots (p_r^{a_r}-p_r^{a_r-1}) \qquad \text{(1)}\\
    &=p_1^{a_1}(1-\frac{1}{p_1})\cdot p_2^{a_2}(1-\frac{1}{p_2})\cdot p_5^{a_5}(1-\frac{1}{p_5})\cdots p_r^{a_r}(1-\frac{1}{p_r})\\
    &= p_1^{a_1}p_2^{a_2}p_5^{a_5}\dots p_r^{a_r}\cdot (1-\frac{1}{p_1})(1-\frac{1}{p_2})(1-\frac{1}{p_5})\cdots (1-\frac{1}{p_r})\\
    &=n\prod_{i=1}^r (1-\frac{1}{p_i}) \qquad \text{(2)}
\end{align*}\qed

\underline{Example:} The only prime divisors of $100$ are $2$ and $5$, so $\phi(100)=100(1-\frac{1}{2})(1-\frac{1}{5})=100\cdot \frac{1}{2}\cdot \frac{4}{5}=40$.

\underline{Example:} The only prime divisors of $120$ are $2,3,5$, so $\phi(120)=120(1-\frac{1}{2})(1-\frac{1}{3})(1-\frac{1}{5})=120\cdot \frac{1}{2}\cdot \frac{2}{3}\cdot \frac{4}{5}=32$.

\underline{Sketch of Proof of Prop 8.6:} Write $\zz_n=\{0,1,\dots,n-1\}$. Then, define the function \begin{align*}
    F:\zz_{mn}&\to \zz_m\times \zz_n\\
\end{align*}by $F(a)=(a\pmod m, a\mod n)$. Then, $F$ is a bijection. Additionally, assuming $\gcd(m,n)=1$, the CRT says that $a\equiv b\mod mn$ if and only if $a\equiv b\mod m$ and $a\equiv b\mod n$. Thus, $F$ restricts to a bijection between the invertibles $\mod mn$ and the pairs of invertibles $\mod m$ and $\mod n$. Thus, $\phi(mn)=\phi(m)\cdot \phi(n)$. \qed 

\underline{Note:} In fact, it's a ring isomorphism because $F(a+b)=(a+b\mod m, a+b\mod n)=F(a)+F(b)$ and $F(ab)=(ab\mod m, ab\mod n)=F(a)F(b)$. 

\end{document}